### 2017-10-11 proposal.tex
first commit

\documentclass{article}
\usepackage[tt=false]{libertine}
\usepackage{hyperref}
\usepackage{fullpage}

\title{Safe Shared Memory for Concurrent Algol}
\author{Farzaneh Derakhshan, Klaas Pruiksma, Jonathan Sterling}

\newcommand\Lang[1]{{\sffamily\bfseries{#1}}}
\newcommand\LangCA{\Lang{CA}}
\newcommand\LangMA{\Lang{MA}}
\newcommand\EOf[4]{{#1}\vdash_{#2}{#3}:{#4}}
\newcommand\MOf[4]{{#1}\vdash_{#2}{#3}\div{#4}}
\newcommand\IsProc[3]{{#1}\vdash_{#2}{#3}\ \mathsf{proc}}
\newcommand\IsAction[3]{{#1}\vdash_{#2}{#3}\ \mathsf{action}}

\begin{document}

\maketitle

\begin{abstract}
  [Please put some abstract material here.]
\end{abstract}

\section{Background on programming langauges}

Farzaneh will write some general amterial here.

\section{Modernized and Concurrent Algol}

Our starting point is the language Concurrent Algol (\LangCA) defined
in Chapter 40 of Harper's \emph{Practical Foundations for Programming
  Languages}~\cite{harper:2016}. \LangCA{} is itself an extension of
Harper's Modernized Algol (\LangMA).

\subsection{Algol-family languages}

Algol was one of the earliest practical imperative programming
languages, and has for many years been considered the quintessential
``clean room'' core language for the scientific study of sequential,
imperative programming~\cite{reynolds:1997}; its distinctive features
has sparked a rich and sophisticated body of developments in
mathematical semantics of programming
languages~\cite{mccusker-power:2010}.

Harper's \LangMA{} differs from traditional presentations of
Algol-family languages in that it surfaces the distinction between
variables, assignables (mutable resources) and assignable
references. In \LangMA{}, a variable is a \emph{mathematical
  variable}, i.e.\ a placeholder for a term; variables are given
meaning through substitution. On the other hand, an assignable is the
name of a location in memory; finally, assignable references are a
device for passing locations in code. By restoring the correct notion
of variable to its rightful place, Harper's presentation clarifies the
structure and semantics of Algol-family languages.

\paragraph{The Essence of \LangMA{}}

The language \LangMA{} is based fundamentally on a modal distinction
between \emph{command} and \emph{expression}. A command is a piece of
code that can ``do'' something, whereas an expression is a piece of
code which calculates a value without any side effects. This is what
sets the Algol family apart from other language families which support
imperative programming (such as \Lang{C}, \Lang{ML}, \Lang{Go},
\Lang{Rust}, etc.).

Accordingly, the static semantics of \LangMA{} are organized around
two forms of assertion:

\begin{enumerate}
\item $\EOf{\Gamma}{\Sigma}{e}{\tau}$ means that $e$ is an \emph{expression}
  of type $\tau$ mentioning variables from $\Gamma$ and assignables
  (memory location names) from $\Sigma$.
\item $\MOf{\Gamma}{\Sigma}{m}{\tau}$ means that $m$ is a
  \emph{command} of type $\tau$ mentioning variables from $\Gamma$ and
  assignables from $\Sigma$.
\end{enumerate}

An operational semantics is given for \LangMA{}, which specifies how
expressions evaluate, and how commands compute relative to a heap
(memory state).

\subsection{Concurrent Algol}

\LangCA{} is an extension of \LangMA{} with a notion of \emph{process}
and their concurrent composition, together with a notion of channel
and message passing. Whilst \LangMA{} had a notion of mutable store,
\LangCA{} does away with this; Harper argues that mutable store can be
eschewed because it can be simulated using message passing.

This simplifying assumption is helpful for the development of a clean
and simple semantics, but it may not be a good idea in practice. While
shared memory comes with all sorts of disadvantages, it can be very
important for the sake of performance. Scientists who engineer
programming languages which are intended to be used in practice must
account for this.

\paragraph{What does \LangCA{} add?}
\LangCA{} adds to \LangMA{}'s static semantics the following forms of
assertion:
\begin{enumerate}
\item $\IsProc{\Gamma}{\Sigma}{p}$ means that $p$ is a \emph{process}
  relative to the variables from $\Gamma$ and the channel names from
  $\Sigma$.
\item $\IsAction{\Gamma}{\Sigma}{\alpha}$ means that $\alpha$ is an
  \emph{action} relative to $\Gamma$, $\Sigma$.
\end{enumerate}

Processes are inductively defined as the \emph{idle process}, the
atomic process $\mathtt{run}(m)$ (which executes the command $m$), and
the concurrent composition of two processes $p_1\otimes p_2$, and
finally the process $\nu\alpha.p$ which generates a fresh
communication channel and proceeds with the process $p$.

Whereas the operational semantics of \LangMA{} included a pure
transition system for expressions and a reduction semantics for
commands relative to a heap or memory state, \LangCA{} adds to this a
\emph{labelled transition system} which describes the evolution of a
system of concurrent processes. Actions $\alpha$ are the labels of
this transition system, and express the ways in which a process can
evolve; actions include querying and signalling along channels.

\paragraph{What do we hope to add to \LangCA{}?}
We hope to add primitives for shared memory to \LangCA{} and show that
our semantics is simulated by the pure message-passing implementation
of shared resources from \LangCA{}. Then, we will extend the critical
safety proofs (for instance, progress for well-typed processes) over
our extended language.

\section{Our Proposal}

Klaas will write this.

\bibliographystyle{abbrv}
\bibliography{bibtex-references/refs}

\end{document}

### 2017-10-11 proposal.tex
...
add a draft abstract

@@
 \begin{abstract}
-  [Please put some abstract material here.]
+  Establishing safety properties for code that mixes concurrency with
+  (shared) memory is notoriously difficult; the experience of the past
+  several years suggests that restricted languages can provide a
+  suitable means to achieve general safety results without sacrificing
+  performance. Languages like \Lang{Rust} and \Lang{Go} are a step in
+  the right direction, but are not currently backed by a precise
+  semantics; we will attempt to thoroughly develop the theory and
+  practice of a small Algol-family language featuring concurrency and
+  shared memory, establishing several crucial safety properties at the
+  language level.
 \end{abstract}
@@
 
-Farzaneh will write some general amterial here.
+Farzaneh will write some general material here.
 

### 2017-10-12 proposal.tex
Added 'Our Proposal' section to proposal. More revisions to that section may follow.

@@
 
-Klaas will write this.
+We plan to create a language building on the ideas of \LangCA{} which has primitive concepts for concurrency, shared memory, and locking.
+Our intent is that this language will disallow writing programs with certain classes of errors ranging from the relatively minor (For example, memory leaks or multiple acquisition of the same lock) to more complex ones including data races (potentially with some restrictions) and deadlocks.
+
+Our 75\% goal is to develop this language and provide evidence that it is both practically usable and avoids errors by writing some sample systems.
+
+Our 100\% goal is to strengthen our justification that the language avoids errors by providing proofs (possibly with some conditions) that those errors cannot occur in a well-typed program.
+While we believe that some of these proofs will be easy to complete, we believe that those for more complex classes of errors will lead us to revise the language in order to get them to work or to work with fewer side conditions, justifying the gap between this and the 75\% goal.
+
+Finally, our 125\% goal is to increase ease-of-use of this language by creating a simplified form for use in writing programs which can then be processed into the more expressive but more complex base language.
+Since many of the conditions surrounding the use of shared memory (or just memory in general) and locks can be inferred from the code as written, we believe that it should be feasible to leave large parts of the memory management and lock management up to an elaboration step that preprocesses the simple program to make memory and lock management explicit.
 

### 2017-10-12 proposal.tex
incorporate slightly revised version of Farzaneh's text

@@
 \usepackage{hyperref}
+\usepackage{mathpartir}
+\usepackage{amsthm}
 \usepackage{fullpage}
@@
 \author{Farzaneh Derakhshan, Klaas Pruiksma, Jonathan Sterling}
+
+\newtheorem{theorem}{Theorem}[section]
+\newtheorem{corollary}{Corollary}[theorem]
+\newtheorem{lemma}[theorem]{Lemma}
 
@@
 
-Farzaneh will write some general material here.
+A proramming language is specified by explaining both its
+\emph{structure} and its \emph{computational meaning}; these are
+captured in the static and the dynamic semantics respectively. Armed
+with such a definition, it is then possible to prove critical
+correctness and safety properties both for a language as a whole, and
+also for individual programs in this language.
+
+One of these basic safety criteria is called ``type safety'', a
+property which is usually factored into \emph{type preservation}
+(programs do not change their type as they execute) and
+\emph{progress} (any well-typed program can take a step of
+computation, or is a value). These laws may seem quite basic, but
+depending on the definition and sophistication of the language, they
+can imply extremely strong semantic properties such as liveness,
+memory safety, causality~\cite{atkey-mcbride:2013}, and even cost or
+complexity bounds.
+
+To give a very simple example of static and dynamic semantics,
+consider the following arithmetic language, adapted
+from~\cite{harper:2016}:
+
+\newcommand\IsOk[1]{{#1}\ \mathsf{ok}}
+\newcommand\IsVal[1]{{#1}\ \mathsf{val}}
+\newcommand\Step[2]{{#1}\mapsto{#2}}
+\newcommand\Plus[2]{\mathtt{plus} (#1, #2)}
+\newcommand\Num[1]{\mathtt{num}[#1]}
+
+\paragraph{Statics} We define the static semantics by specifying which
+expressions are ``OK'' (well-typed):
+\begin{mathpar}
+  \inferrule{
+  }{
+    \IsOk{\Num{n}}
+  }
+  \and
+  \inferrule{
+    \IsOk{e_1}
+    \\
+    \IsOk{e_2}
+  }{
+    \IsOk{\Plus{e_1}{e_2}}
+  }
+\end{mathpar}
+
+In a more sophisticated language, the static semantics would be given
+by a much more complicated form of judgment than $\IsOk{e}$, as we
+will see later on.
+
+\paragraph{Dynamics}
+The dynamics are given by a transition system, writing $\IsVal{e}$ to
+mean that $e$ is a \emph{value}, and $\Step{e}{e'}$ to mean that the
+program $e$ takes a step to $e'$.
+\begin{mathpar}
+  \inferrule{
+  }{
+    \IsVal{\Num{n}}
+  }
+  \and
+  \inferrule{
+    \Step{e_1}{e_1'}
+  }{
+    \Step{\Plus{e_1}{e_2}}{\Plus{e_1'}{e_2}}
+  }
+  \and
+  \inferrule{
+    \IsVal{e_1}
+    \\
+    \Step{e_2}{e_2'}
+  }{
+    \Step{\Plus{e_1}{e_2}}{\Plus{e_1}{e_2'}}
+  }
+  \and
+  \inferrule{
+    n_1 + n_2 = n_3
+  }{
+    \Step{
+      \Plus{\Num{n_1}}{\Num{n_2}}
+    }{
+      \Num{n_3}
+    }
+  }
+\end{mathpar}
+
+At this point, we can prove the type safety theorem in two lemmas:
+\begin{lemma}[Progress]
+  If $\IsOk{e}$, then either $\IsVal{e}$ or $\Step{e}{e'}$ for some
+  $e'$.
+\end{lemma}
+\begin{lemma}[Preservation]
+  If $\IsOk{e}$ and $\Step{e}{e'}$, then $\IsOk{e'}$.
+\end{lemma}
+
+Of course, this was a very trivial example, but this methodology
+scales up to problems of seemingly arbitrary complexity. For instance,
+a type system inspired by Kripke semantics and the modal logic S5 has
+been deployed in the ML5 language for distributed computing, enabling
+one to develop distributed applications within a single unified
+program~\cite{murphy-crary-harper:2008}. Concurrency primitives have
+been added to functional languages such as
+Facile~\cite{giacolone:1989}, Concurrent ML~\cite{reppy:1991,reppy:2007},
+JOCaml~\cite{conchon-fessant:1999}, Concurrent
+Haskell~\cite{jones-gordon-finne:1996} and Concurrent
+Algol~\cite{harper:2016}. In another work, Balzer and Pfenning
+augmented the session-typed linear logic with shared memory, using
+shared channels and shared memory with an acquire-release discipline.
 

### 2017-10-12 proposal.tex
more citation

@@
 \usepackage{fullpage}
+
+\usepackage{sectsty}
+\allsectionsfont{\sffamily}
 
@@
 
-\section{Background on programming langauges}
+\section{What are languages good for?}
 
@@
 been added to functional languages such as
-Facile~\cite{giacolone:1989}, Concurrent ML~\cite{reppy:1991,reppy:2007},
-JOCaml~\cite{conchon-fessant:1999}, Concurrent
-Haskell~\cite{jones-gordon-finne:1996} and Concurrent
+Facile~\cite{giacolone:1989}, Concurrent
+ML~\cite{reppy:1991,reppy:2007}, JOCaml~\cite{conchon-fessant:1999},
+Concurrent Haskell~\cite{jones-gordon-finne:1996},
+Rust~\cite{turon:2012}, Go~\cite{donovan:2015} and Concurrent
 Algol~\cite{harper:2016}. In another work, Balzer and Pfenning

### 2017-10-12 proposal.tex
typo

@@
 We plan to create a language building on the ideas of \LangCA{} which has primitive concepts for concurrency, shared memory, and locking.
-Our intent is that this language will disallow writing programs with certain classes of errors ranging from the relatively minor (For example, memory leaks or multiple acquisition of the same lock) to more complex ones including data races (potentially with some restrictions) and deadlocks.
+Our intent is that this language will disallow writing programs with certain classes of errors ranging from the relatively minor (for example, memory leaks or multiple acquisition of the same lock) to more complex ones including data races (potentially with some restrictions) and deadlocks.
 

### 2017-10-12 proposal.tex
fix missing reference

@@
 augmented the session-typed linear logic with shared memory, using
-shared channels and shared memory with an acquire-release discipline.
+shared channels and shared memory with an acquire-release
+discipline~\cite{balzer-pfenning:2017}.
 

### 2017-11-04 proposal.tex
...

@@
 
+We believe that each of these is a feasible goal, with the bulk of the work lying in the design and creation of the language, leading us to choose this as the 75\% goal --- without a complete language and some examples, we do not have a complete project.
+However, we believe that the language and examples would suffice as evidence for the feasibility of our goals, and would improve on existing languages by giving formal definitions which allow for the desired properties to be proven, albeit possibly with some modifications to the language.
+Despite this, the proofs in our 100\% goal are much stronger evidence, and make it much more feasible to evaluate the utility of this project.
+We include this as our 100\% goal because it will yield a usable language with provable guarantees, which is what we set out to do in the first place.
+There is little to be done from a correctness perspective past this other than potentially extending to cover more sorts of errors, and so from this, we determined that it made more sense to set our 125\% goal in the direction of increased usability, as described above.
+This is the most concretely-evaluable part of the work, as it allows us to make comparisons such as the amount of our code examples that can be automatically generated, which will serve as a proxy for the ease of writing code in this simplified system compared to the original system.
+
 \bibliographystyle{abbrv}

### 2017-11-15 src/network_c/c0_network.h
Networking headers added.

@@ -1,0 +1,37 @@
+#include <poll.h>
+#include <unistd.h>
+
+#include <sys/types.h>
+#include <sys/socket.h>
+
+int32_t c0_dgram_socket ();
+
+
+/**
+ *  Socket I/O
+ */
+
+int32_t c0_recv_msglen (int32_t sockfd);
+
+// src_sin_family, src_sin_port, src_sin_addr are all returned. Likely only
+// the last two are necessary, but we keep family just in case.
+int32_t c0_recvfrom (int32_t sockfd, int32_t* buf, int32_t len,
+                     int32_t* src_sin_family, int32_t* src_sin_port,
+                     int32_t* src_sin_addr);
+
+// ssize_t sendto(int sockfd, const void *buf, size_t len, int flags,
+//                const struct sockaddr *dest_addr, socklen_t addrlen);
+
+int32_t c0_sendto (int32_t sockfd, int32_t* buf, int32_t len,
+                   int32_t dst_sin_family, int32_t dst_sin_port,
+                   int32_t dst_sin_addr);
+
+
+/**
+ *  Polling sockets for readiness.
+ */
+
+// TODO: Timeouts needed?
+int32_t c0_poll_read (int32_t sockfd, int32_t timeout);
+
+int32_t c0_poll_write (int32_t sockfd, int32_t timeout);

### 2017-11-16 proposal.tex
proposal: add outline

--- 2017-10-12 proposal.tex
@@
 
-\title{Safe Shared Memory for Concurrent Algol}
+\title{Implementing DNS in C0}
 \author{Farzaneh Derakhshan, Klaas Pruiksma, Jonathan Sterling}
@@
 \begin{abstract}
-  Establishing safety properties for code that mixes concurrency with
-  (shared) memory is notoriously difficult; the experience of the past
-  several years suggests that restricted languages can provide a
-  suitable means to achieve general safety results without sacrificing
-  performance. Languages like \Lang{Rust} and \Lang{Go} are a step in
-  the right direction, but are not currently backed by a precise
-  semantics; we will attempt to thoroughly develop the theory and
-  practice of a small Algol-family language featuring concurrency and
-  shared memory, establishing several crucial safety properties at the
-  language level.
+  Please write an abstract.
 \end{abstract}
 
-\section{What are languages good for?}
+\section{Background and Related Work}
 
-A proramming language is specified by explaining both its
-\emph{structure} and its \emph{computational meaning}; these are
-captured in the static and the dynamic semantics respectively. Armed
-with such a definition, it is then possible to prove critical
-correctness and safety properties both for a language as a whole, and
-also for individual programs in this language.
+\subsection{What is DNS?}
 
-One of these basic safety criteria is called ``type safety'', a
-property which is usually factored into \emph{type preservation}
-(programs do not change their type as they execute) and
-\emph{progress} (any well-typed program can take a step of
-computation, or is a value). These laws may seem quite basic, but
-depending on the definition and sophistication of the language, they
-can imply extremely strong semantic properties such as liveness,
-memory safety, causality~\cite{atkey-mcbride:2013}, and even cost or
-complexity bounds.
+Jon
 
-To give a very simple example of static and dynamic semantics,
-consider the following arithmetic language, adapted
-from~\cite{harper:2016}:
+\subsection{C0 and Concurrent C0}
 
-\newcommand\IsOk[1]{{#1}\ \mathsf{ok}}
-\newcommand\IsVal[1]{{#1}\ \mathsf{val}}
-\newcommand\Step[2]{{#1}\mapsto{#2}}
-\newcommand\Plus[2]{\mathtt{plus} (#1, #2)}
-\newcommand\Num[1]{\mathtt{num}[#1]}
+Farzaneh.
 
-\paragraph{Statics} We define the static semantics by specifying which
-expressions are ``OK'' (well-typed):
-\begin{mathpar}
-  \inferrule{
-  }{
-    \IsOk{\Num{n}}
-  }
-  \and
-  \inferrule{
-    \IsOk{e_1}
-    \\
-    \IsOk{e_2}
-  }{
-    \IsOk{\Plus{e_1}{e_2}}
-  }
-\end{mathpar}
+\subsection{Related Work}
+Jon.
 
-In a more sophisticated language, the static semantics would be given
-by a much more complicated form of judgment than $\IsOk{e}$, as we
-will see later on.
+\section{Our Project}
 
-\paragraph{Dynamics}
-The dynamics are given by a transition system, writing $\IsVal{e}$ to
-mean that $e$ is a \emph{value}, and $\Step{e}{e'}$ to mean that the
-program $e$ takes a step to $e'$.
-\begin{mathpar}
-  \inferrule{
-  }{
-    \IsVal{\Num{n}}
-  }
-  \and
-  \inferrule{
-    \Step{e_1}{e_1'}
-  }{
-    \Step{\Plus{e_1}{e_2}}{\Plus{e_1'}{e_2}}
-  }
-  \and
-  \inferrule{
-    \IsVal{e_1}
-    \\
-    \Step{e_2}{e_2'}
-  }{
-    \Step{\Plus{e_1}{e_2}}{\Plus{e_1}{e_2'}}
-  }
-  \and
-  \inferrule{
-    n_1 + n_2 = n_3
-  }{
-    \Step{
-      \Plus{\Num{n_1}}{\Num{n_2}}
-    }{
-      \Num{n_3}
-    }
-  }
-\end{mathpar}
+\subsection{Network interface}
 
-At this point, we can prove the type safety theorem in two lemmas:
-\begin{lemma}[Progress]
-  If $\IsOk{e}$, then either $\IsVal{e}$ or $\Step{e}{e'}$ for some
-  $e'$.
-\end{lemma}
-\begin{lemma}[Preservation]
-  If $\IsOk{e}$ and $\Step{e}{e'}$, then $\IsOk{e'}$.
-\end{lemma}
+Klaas.
 
-Of course, this was a very trivial example, but this methodology
-scales up to problems of seemingly arbitrary complexity. For instance,
-a type system inspired by Kripke semantics and the modal logic S5 has
-been deployed in the ML5 language for distributed computing, enabling
-one to develop distributed applications within a single unified
-program~\cite{murphy-crary-harper:2008}. Concurrency primitives have
-been added to functional languages such as
-Facile~\cite{giacolone:1989}, Concurrent
-ML~\cite{reppy:1991,reppy:2007}, JOCaml~\cite{conchon-fessant:1999},
-Concurrent Haskell~\cite{jones-gordon-finne:1996},
-Rust~\cite{turon:2012}, Go~\cite{donovan:2015} and Concurrent
-Algol~\cite{harper:2016}. In another work, Balzer and Pfenning
-augmented the session-typed linear logic with shared memory, using
-shared channels and shared memory with an acquire-release discipline.
+\subsection{Marshalling and unmarshalling}
 
-\section{Modernized and Concurrent Algol}
+Jon
 
-Our starting point is the language Concurrent Algol (\LangCA) defined
-in Chapter 40 of Harper's \emph{Practical Foundations for Programming
-  Languages}~\cite{harper:2016}. \LangCA{} is itself an extension of
-Harper's Modernized Algol (\LangMA).
+\section{Evaluation Plan}
 
-\subsection{Algol-family languages}
-
-Algol was one of the earliest practical imperative programming
-languages, and has for many years been considered the quintessential
-``clean room'' core language for the scientific study of sequential,
-imperative programming~\cite{reynolds:1997}; its distinctive features
-has sparked a rich and sophisticated body of developments in
-mathematical semantics of programming
-languages~\cite{mccusker-power:2010}.
-
-Harper's \LangMA{} differs from traditional presentations of
-Algol-family languages in that it surfaces the distinction between
-variables, assignables (mutable resources) and assignable
-references. In \LangMA{}, a variable is a \emph{mathematical
-  variable}, i.e.\ a placeholder for a term; variables are given
-meaning through substitution. On the other hand, an assignable is the
-name of a location in memory; finally, assignable references are a
-device for passing locations in code. By restoring the correct notion
-of variable to its rightful place, Harper's presentation clarifies the
-structure and semantics of Algol-family languages.
-
-\paragraph{The Essence of \LangMA{}}
-
-The language \LangMA{} is based fundamentally on a modal distinction
-between \emph{command} and \emph{expression}. A command is a piece of
-code that can ``do'' something, whereas an expression is a piece of
-code which calculates a value without any side effects. This is what
-sets the Algol family apart from other language families which support
-imperative programming (such as \Lang{C}, \Lang{ML}, \Lang{Go},
-\Lang{Rust}, etc.).
-
-Accordingly, the static semantics of \LangMA{} are organized around
-two forms of assertion:
-
-\begin{enumerate}
-\item $\EOf{\Gamma}{\Sigma}{e}{\tau}$ means that $e$ is an \emph{expression}
-  of type $\tau$ mentioning variables from $\Gamma$ and assignables
-  (memory location names) from $\Sigma$.
-\item $\MOf{\Gamma}{\Sigma}{m}{\tau}$ means that $m$ is a
-  \emph{command} of type $\tau$ mentioning variables from $\Gamma$ and
-  assignables from $\Sigma$.
-\end{enumerate}
-
-An operational semantics is given for \LangMA{}, which specifies how
-expressions evaluate, and how commands compute relative to a heap
-(memory state).
-
-\subsection{Concurrent Algol}
-
-\LangCA{} is an extension of \LangMA{} with a notion of \emph{process}
-and their concurrent composition, together with a notion of channel
-and message passing. Whilst \LangMA{} had a notion of mutable store,
-\LangCA{} does away with this; Harper argues that mutable store can be
-eschewed because it can be simulated using message passing.
-
-This simplifying assumption is helpful for the development of a clean
-and simple semantics, but it may not be a good idea in practice. While
-shared memory comes with all sorts of disadvantages, it can be very
-important for the sake of performance. Scientists who engineer
-programming languages which are intended to be used in practice must
-account for this.
-
-\paragraph{What does \LangCA{} add?}
-\LangCA{} adds to \LangMA{}'s static semantics the following forms of
-assertion:
-\begin{enumerate}
-\item $\IsProc{\Gamma}{\Sigma}{p}$ means that $p$ is a \emph{process}
-  relative to the variables from $\Gamma$ and the channel names from
-  $\Sigma$.
-\item $\IsAction{\Gamma}{\Sigma}{\alpha}$ means that $\alpha$ is an
-  \emph{action} relative to $\Gamma$, $\Sigma$.
-\end{enumerate}
-
-Processes are inductively defined as the \emph{idle process}, the
-atomic process $\mathtt{run}(m)$ (which executes the command $m$), and
-the concurrent composition of two processes $p_1\otimes p_2$, and
-finally the process $\nu\alpha.p$ which generates a fresh
-communication channel and proceeds with the process $p$.
-
-Whereas the operational semantics of \LangMA{} included a pure
-transition system for expressions and a reduction semantics for
-commands relative to a heap or memory state, \LangCA{} adds to this a
-\emph{labelled transition system} which describes the evolution of a
-system of concurrent processes. Actions $\alpha$ are the labels of
-this transition system, and express the ways in which a process can
-evolve; actions include querying and signalling along channels.
-
-\paragraph{What do we hope to add to \LangCA{}?}
-We hope to add primitives for shared memory to \LangCA{} and show that
-our semantics is simulated by the pure message-passing implementation
-of shared resources from \LangCA{}. Then, we will extend the critical
-safety proofs (for instance, progress for well-typed processes) over
-our extended language.
-
-\section{Our Proposal}
-
-We plan to create a language building on the ideas of \LangCA{} which has primitive concepts for concurrency, shared memory, and locking.
-Our intent is that this language will disallow writing programs with certain classes of errors ranging from the relatively minor (for example, memory leaks or multiple acquisition of the same lock) to more complex ones including data races (potentially with some restrictions) and deadlocks.
-
-Our 75\% goal is to develop this language and provide evidence that it is both practically usable and avoids errors by writing some sample systems.
-
-Our 100\% goal is to strengthen our justification that the language avoids errors by providing proofs (possibly with some conditions) that those errors cannot occur in a well-typed program.
-While we believe that some of these proofs will be easy to complete, we believe that those for more complex classes of errors will lead us to revise the language in order to get them to work or to work with fewer side conditions, justifying the gap between this and the 75\% goal.
-
-Finally, our 125\% goal is to increase ease-of-use of this language by creating a simplified form for use in writing programs which can then be processed into the more expressive but more complex base language.
-Since many of the conditions surrounding the use of shared memory (or just memory in general) and locks can be inferred from the code as written, we believe that it should be feasible to leave large parts of the memory management and lock management up to an elaboration step that preprocesses the simple program to make memory and lock management explicit.
+Klaas.
 

### 2017-11-16 proposal.tex
start filling out material on marshalling

@@
 \usepackage[tt=false]{libertine}
-\usepackage{hyperref}
 \usepackage{mathpartir}
 \usepackage{amsthm}
-\usepackage{fullpage}
+\usepackage[colorlinks=true]{hyperref}
+\usepackage{bytefield}
+%\usepackage{fullpage}
+\usepackage{listings}
+
+\lstset{language=C}
+\lstset{basicstyle=\ttfamily, columns=flexible}
 
@@
 
-\subsection{C0 and Concurrent C0}
+\subsection{C0 and Concurrent C0}\label{sec:c0}
 
@@
 
-\subsection{Network interface}
+Our original project proposal was to develop a language for safe
+concurrent programming with precise formal guarantees; however, the
+negotiation process led us to choose a completely unrelated project,
+which we describe in this section. We will be developing an
+implementation of the DNS protocol in the memory-safe \Lang{C0}
+language (see \S\ref{sec:c0}); we have divided our work into two
+principle components.
+
+\subsection{Network interface}\label{sec:network-interface}
 
@@
 
-Jon
+Unlike \Lang{C}, the \Lang{C0} language has only a single type of
+integers, which are represented internally as 32-bit integers. The
+network interface (see \S\ref{sec:network-interface}) will deal with
+requests and responses formatted as arrays of 32-bit integers; it is
+the responsibility of the marshalling-and-unmarshalling module to
+unpack this data into structured representations of core DNS
+datatypes.
+
+\begin{figure}
+  \centering
+  \begin{bytefield}[bitwidth=1.5em]{16}
+    \wordbox{1}{\texttt{ID}}
+    \\
+    \bitbox{1}{\texttt{QR}}
+    \bitbox{4}{\texttt{Opcode}}
+    \bitbox{1}{\texttt{AA}}
+    \bitbox{1}{\texttt{TC}}
+    \bitbox{1}{\texttt{RD}}
+    \bitbox{1}{\texttt{RA}}
+    \bitbox{3}{\texttt{Z}}
+    \bitbox{4}{\texttt{RCODE}}
+    \\
+    \wordbox{1}{\texttt{QDCOUNT}}
+    \\
+    \wordbox{1}{\texttt{ANCOUNT}}
+    \\
+    \wordbox{1}{\texttt{NSCOUNT}}
+    \\
+    \wordbox{1}{\texttt{ARCOUNT}}
+  \end{bytefield}
+  \caption{The layout specification of message headers in
+    DNS.}\label{fig:layout-message-header}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \begin{bytefield}{16}
+    \wordbox[lrt]{1}{\texttt{NAME}}\\
+    \skippedwords\\
+    \wordbox[lrb]{1}{}\\
+    \wordbox{1}{\texttt{TYPE}}\\
+    \wordbox{1}{\texttt{CLASS}}\\
+    \wordbox{2}{\texttt{TTL}}\\
+    \wordbox[lrt]{1}{\texttt{RDATA}}\\
+    \skippedwords\\
+    \wordbox[lrb]{1}{}
+  \end{bytefield}
+  \caption{The bit layout specification of resource records in
+    DNS.}\label{fig:layout-resource-record}
+\end{figure}
+
+\begin{figure}
+  \begin{lstlisting}[frame=single]
+struct header {
+  int id;
+  int qr;
+  int opcode;
+  int aa;
+  int tc;
+  int rd;
+  int ra;
+  int z;
+  int rcode;
+  int qdcount;
+  int ancount;
+  int nscount;
+  int arcount;
+};
+
+typedef struct header* header_t;
+  \end{lstlisting}
+
+  \begin{lstlisting}[frame=single]
+struct domain_name {
+  string label;
+  struct domain_name* next;
+};
+
+typedef struct domain_name* domain_name_t;
+
+struct resource_record {
+  domain_name_t name;
+  int type;
+  int class;
+  int ttl;
+  int rd_length;
+  int[] rdata;
+};
+
+typedef struct resource_record* resource_record_t;
+  \end{lstlisting}
+  \caption{Example DNS data structures in our \Lang{C0} implementation.}\label{fig:c0-data-structures}
+\end{figure}
+
+For instance, Figure~\ref{fig:layout-message-header} shows the
+bit-level layout of DNS message headers, and
+Figure~\ref{fig:layout-resource-record} shows the same for resource
+records; Figure~\ref{fig:c0-data-structures} shows how we code these
+structures in our \Lang{C0} implementation. To convert between these
+structured representations and the sequences of words which are used
+at the boundaries of the network
+interface~(\S\ref{sec:network-interface}), we implement functions of
+the following kind:
+
+\begin{lstlisting}
+int parse_domain_name(int cursor, int[] msg, domain_name_t dest);
+int parse_resource_record(int cursor, int[] msg, resource_record_t dest);
+int parse_header(int cursor, int[] msg, header_t dest);
+\end{lstlisting}
+
+These functions take as an argument a cursor (an index in bits into
+the message), an array of 32-bit words (the entire input message), and
+pointer to a structure into which to write the parsed data; they
+return a new cursor. It is important for the modularity of our
+development that cursors represent not indices into the 32-bit cells
+of the message, but rather bit-indices; this is because specific parts
+of DNS data structures may only be a few bits long, so their parsers
+cannot move the cursor forward a full word's length.
+%
+A sample of our parsing code can be found in the supplementary
+appendix to this proposal (\S\ref{appendix:decompression}). Of
+course, we will also implement the inverses to the above functions.
 
@@
 
+\clearpage
+\appendix
+\section{Decompressing Domain Names}\label{appendix:decompression}
+
+\begin{lstlisting}
+int parse_domain_name(int cursor, int[] data, domain_name* dest) {
+  int* len = alloc(int);
+
+  int orig_cursor = cursor;
+  cursor = read_bits(cursor, data, 8, len);
+
+  if (*len == 0) {
+    // case: done
+    *dest = domain_nil();
+    return cursor;
+  }
+
+  if (*len >= 0xc0) {
+    // case: pointer
+    int* ptr = alloc(int);
+    // read the remainder of the first 16 bits as a pointer to an offset in octets
+    cursor = read_bits(orig_cursor + 2, data, 14, ptr);
+    // read the last label of the domain name from the pointer:
+    parse_domain_name(*ptr * 8, data, dest);
+    return cursor;
+  }
+
+  // case: ordinary label
+  // now '*len' specifies the number of octets in the label
+  char[] chars = alloc_array(char, *len);
+  for (int i = 0; i < *len; i++) {
+    int* octet = alloc(int);
+    cursor = read_bits(cursor, data, 8, octet);
+    chars[i] = char_chr(*octet);
+  }
+
+  string lbl = string_from_chararray(chars);
+
+  *dest = domain_cons(lbl, *dest);
+  return parse_domain_name(cursor, data, dest);
+}
+\end{lstlisting}
+
 \end{document}

### 2017-11-16 proposal.tex
add project.bib for local bibliographic entries
add some bibliography entries
discuss related work

@@
 \usepackage{listings}
+\usepackage{microtype}
 
@@
 
-\newcommand\Lang[1]{{\sffamily\bfseries{#1}}}
-\newcommand\LangCA{\Lang{CA}}
-\newcommand\LangMA{\Lang{MA}}
+\newcommand\Kwd[1]{{\sffamily\bfseries{#1}}}
 \newcommand\EOf[4]{{#1}\vdash_{#2}{#3}:{#4}}
@@
 \subsection{Related Work}
-Jon.
+
+\paragraph{The \Kwd{Fox} Project}
+An earlier example of an implementation of the DNS protocol in a safe
+and formally specified language is given by the \Kwd{Fox} project,
+undertaken at Carnegie Mellon University in the
+1990s~\cite{biagioni-harper-lee-milnes:1994,
+  biagioni-harper-lee:2001}. Their principal artifact was a library
+called \Kwd{FoxNet} which implemented an array of low-level network
+infrastructure, including \Kwd{TCP}, \Kwd{IP}, \Kwd{ARP} and \Kwd{DNS}
+in the \Kwd{Standard ML} language.
+
+Unlike our development, \Kwd{FoxNet} had a highly modular character
+which was enabled by the host language's sophisticated module system
+(at the time, the only language to support signatures, structures and
+generative functors, which are crucial for safe abstraction).
+
+\paragraph{Other implementations} More recently, safe languages like
+\Kwd{Rust} and \Kwd{Go} have been gaining popularity. Both languages
+have hosted implementations of the \Kwd{DNS}
+protocol~\cite{github:trust-dns,github:miekg-dns}
 
@@
 which we describe in this section. We will be developing an
-implementation of the DNS protocol in the memory-safe \Lang{C0}
+implementation of the DNS protocol in the memory-safe \Kwd{C0}
 language (see \S\ref{sec:c0}); we have divided our work into two
@@
 
-Unlike \Lang{C}, the \Lang{C0} language has only a single type of
+Unlike \Kwd{C}, the \Kwd{C0} language has only a single type of
 integers, which are represented internally as 32-bit integers. The
@@
   \end{lstlisting}
-  \caption{Example DNS data structures in our \Lang{C0} implementation.}\label{fig:c0-data-structures}
+  \caption{Example DNS data structures in our \Kwd{C0} implementation.}\label{fig:c0-data-structures}
 \end{figure}
@@
 records; Figure~\ref{fig:c0-data-structures} shows how we code these
-structures in our \Lang{C0} implementation. To convert between these
+structures in our \Kwd{C0} implementation. To convert between these
 structured representations and the sequences of words which are used
@@
 
+\nocite{willsey-prabhu-pfenning:2016, rfc:1034, rfc:1035}
 \bibliographystyle{abbrv}
-\bibliography{bibtex-references/refs}
+\bibliography{bibtex-references/refs,project}
 

### 2017-11-17 proposal.tex
update outline
add Farzaneh's contribution

@@
 
-Jon
+Who will write this? I can write some, but I don't really know very much. -Jon
 
@@
 
-Farzaneh.
+The \Kwd{C0} programming language is a restricted subset of \Kwd{C}
+that provides implicit memory safety to
+programmers~\cite{pfenning:c0-reference}. Restrictions on pointer
+arithmetic and casting, for example, allow the language to detect and
+prevent any attempt to access an array out of its bounds. Furthermore,
+there is no need for error-prone manual memory management, since
+\Kwd{C0} is itself garbage-collected. Further safety conditions can be
+enforced with contracts, which are a class of assertions including
+loop invariants, pre- and post-conditions, and dynamically checkable
+correctness conditions for programs.
+
+\Kwd{Concurrent C0} (\Kwd{CC0}) is an extension of \Kwd{C0} that
+offers concurrency based on message passing by providing processes as
+the units of concurrent execution and channels as their message
+buffers~\cite{willsey-prabhu-pfenning:2016}. Concurrent processes
+communicate over channels using a protocol that is disciplined by
+session types~\cite{caires-pfenning-toninho:2013,
+  balzer-pfenning:2017}. In this protocol, each channel is typed such
+that it can either send or receive a message and, thus, it
+deliberately models the client/provider relation.
+
+However, the protocol is also able to characterize processes that are
+not completely sequential, and branch into different sequences as they
+encounter choices. This typing system, also, guarantees that data is
+send and received in the correct order and by the correct processes
+(session fidelity), and that the communication is always one-to-one.
+The \Kwd{CC0} programming language is secure in terms of both memory
+and communication safety.
 
@@
 
-\nocite{willsey-prabhu-pfenning:2016, rfc:1034, rfc:1035}
+\nocite{rfc:1034, rfc:1035}
 \bibliographystyle{abbrv}

### 2017-11-17 proposal.tex
minor wording changes

@@
 not completely sequential, and branch into different sequences as they
-encounter choices. This typing system, also, guarantees that data is
-send and received in the correct order and by the correct processes
-(session fidelity), and that the communication is always one-to-one.
-The \Kwd{CC0} programming language is secure in terms of both memory
-and communication safety.
+encounter choices. This typing discipline also guarantees that data is
+sent and received in the correct order and by the correct processes
+(an invariant called \emph{session fidelity}), and that the
+communication is always one-to-one.
 

### 2017-11-17 proposal.tex
fix a few typos

@@
 \begin{lstlisting}
-int parse_domain_name(int cursor, int[] msg, domain_name_t dest);
+int parse_domain_name(int cursor, int[] msg, domain_name_t* dest);
 int parse_resource_record(int cursor, int[] msg, resource_record_t dest);
@@
 \begin{lstlisting}
-int parse_domain_name(int cursor, int[] data, domain_name* dest) {
+int parse_domain_name(int cursor, int[] data, domain_name_t* dest) {
   int* len = alloc(int);

### 2017-11-17 proposal.tex
Added the network interface section and a label for the goals/progress section and fixed some typos.

@@
 \title{Implementing DNS in C0}
-\author{Farzaneh Derakhshan, Klaas Pruiksma, Jonathan Sterling}
+\author{Farzaneh Derakhshan \and Klaas Pruiksma \and Jonathan Sterling}
 
@@
 language (see \S\ref{sec:c0}); we have divided our work into two
-principle components.
+principal components.
 
@@
 
-Klaas.
+In its current state, \Kwd{C0} has a very limited standard library.
+In particular, while there are libraries for input and output to files and to console, there is no networking library.
+It may be possible to implement networking using file I/O, but we do not believe that this is reasonable.
+
+As such, we will write a networking library for \Kwd{C0}.
+Due to limitations of the language, however, this library cannot be written entirely in \Kwd{C0} itself.
+Instead, we will write a collection of small functions in \Kwd{C} which act as an interface between \Kwd{C0} code and the \Kwd{C} sockets library.
+On top of this, we can then write the \Kwd{C0} portion of the library, which will provide a higher-level interface for network communication than that provided by the \Kwd{C} portion of the library.
+
+By splitting the library into these two parts, we can keep the \Kwd{C} portion of our codebase limited in scope.
+In particular, we plan to ensure that we never dynamically allocate memory from \Kwd{C} and that any accesses to \Kwd{C0}-allocated memory are done so safely.
+The former is easily checked for, simply by checking that there are no calls to malloc or similar functions, while the latter can be achieved (but less easily checked) by using a collection of functions for safe memory access provided to libraries by the \Kwd{C0} runtime.
+As a result of this and the small size of the \Kwd{C} portion of the codebase, we can easily check all of our code for memory safety, ensuring that any memory safety violations occur outside of our project (e.g.
+in the \Kwd{C} sockets library or in the OS).
+
+Since this project is limited in scope to \Kwd{DNS}, we will not expose the full functionality of the sockets library in our network interface.
+We will instead focus on the functions that create and work with datagram-based sockets, as all of the network communication needed for \Kwd{DNS} is done via the datagram protocol \Kwd{UDP}.
+Additionally, to avoid busy-waiting, we will provide a limited interface to the \texttt{poll} function of the \Kwd{C} standard library in much the same way.
+We believe that this is sufficient to implement our nameserver, but additional library functionality can be added using these methods if necessary.
 
@@
 
+\section{Goals and Progress}
+
 \section{Evaluation Plan}

### 2017-11-17 proposal.tex
turn on fullpage; update outline
Updated network interface section for clarity.

@@
 The former is easily checked for, simply by checking that there are no calls to malloc or similar functions, while the latter can be achieved (but less easily checked) by using a collection of functions for safe memory access provided to libraries by the \Kwd{C0} runtime.
-As a result of this and the small size of the \Kwd{C} portion of the codebase, we can easily check all of our code for memory safety, ensuring that any memory safety violations occur outside of our project (e.g.
+As a result of this and the small size of the \Kwd{C} portion of the codebase, we can easily check all of the \Kwd{C} code for memory safety.
+Combining this with \Kwd{C0}'s memory safety, we ensure that any memory safety violations occur outside of our project (e.g.
 in the \Kwd{C} sockets library or in the OS).

### 2017-11-17 proposal.tex
Goals section finished.

@@
 \usepackage{bytefield}
-%\usepackage{fullpage}
+\usepackage{fullpage}
 \usepackage{listings}
@@
 
-Who will write this? I can write some, but I don't really know very much. -Jon
+Farzaneh
 
@@
 
-\section{Our Project}
+\section{Our Project}\label{sec:our-project}
 
@@
 Since this project is limited in scope to \Kwd{DNS}, we will not expose the full functionality of the sockets library in our network interface.
-We will instead focus on the functions that create and work with datagram-based sockets, as all of the network communication needed for \Kwd{DNS} is done via the datagram protocol \Kwd{UDP}.
+We will instead focus on the functions that create and work with datagram-based sockets, as most of the network communication needed for \Kwd{DNS} is done via the datagram protocol \Kwd{UDP}.
 Additionally, to avoid busy-waiting, we will provide a limited interface to the \texttt{poll} function of the \Kwd{C} standard library in much the same way.
@@
 
+\subsection{75\%}
+
+Our 75\% goal is to develop a basic \Kwd{DNS} nameserver in \Kwd{C0}.
+In order to reduce the scope of this project, we restrict the types of resource data (and by extension, the types of queries) that we handle, and we will only provide support for \Kwd{UDP}-based queries.
+These two restrictions will simplify implementation, but leave enough functionality that we can evaluate our implementation both for correctness and for performance relative to other DNS implementations, as long as we restrict the test queries to those we have implemented.
+
+The bulk of this goal falls into the two components described in \S\ref{sec:our-project}, along with a small ``main'' component that uses these to actually handle queries.
+Currently, we have unmarshalling functions and about a third of the network library implemented.
+We expect that the remainder of this goal should go fairly quickly --- marshalling is no harder than unmarshalling, and the network interface is relatively straightforward to implement once we have decided on what functions are necessary and what their arguments should be.
+
+\subsection{100\%}
+
+To extend the above to a 100\% goal, we plan to extend the functionality of our nameserver by adding caching of requests and potentially adding to the types of resource data we support.
+While a nameserver without caching is not particularly useful, we believe that the portion of code most likely to contain memory errors is in some combination of reading data from the network and parsing that data --- in particular, dealing with message compression.
+As such, the 75\% goal is a proof of concept that demonstrates that these portions can be implemented safely, while the 100\% goal extends that proof of concept to be more useable (though still more limited than commercially available nameservers).
+We expect that in going from our 75\% version to the 100\% version, performance (and the scope of allowable queries) will increase and that we will be able to leave our arguments for correctness unchanged.
+
+\subsection{125\%}
+
+We have two potential 125\% goals, depending on the outcome of our 100\% goal.
+If our nameserver has good performance relative to its competitors, we will work to provide better justification for the correctness of our nameserver, ideally arriving at a proof that statically verifies that all of our memory accesses are safe.
+In doing this, we will be able to safely turn off bounds checking when running our nameserver, increasing performance through verifying correctness.
+If instead our performance is very poor (or not helped much by turning off bounds checking), we will work to optimize our nameserver based on the results of performance testing.
+It is difficult to say what specifically we will do, but we plan to take the general approach of using software like \texttt{perf} to determine what portions of our code are bottlenecks, and trying to make those paths faster.
+
 \section{Evaluation Plan}

### 2017-11-17 proposal.tex
clarify
Evaluation section added.

@@
 The bulk of this goal falls into the two components described in \S\ref{sec:our-project}, along with a small ``main'' component that uses these to actually handle queries.
-Currently, we have unmarshalling functions and about a third of the network library implemented.
+Currently, we have some unmarshalling functions and about a third of the network library implemented.
 We expect that the remainder of this goal should go fairly quickly --- marshalling is no harder than unmarshalling, and the network interface is relatively straightforward to implement once we have decided on what functions are necessary and what their arguments should be.
@@
 
-Klaas.
+Our evaluation breaks down into two parts --- one on the safety, and one on performance.
+As safety is a primary goal of this project, we plan to evaluate our code for safety.
+A large part of this will come for free from the memory safety guarantees of \Kwd{C0}, with the rest coming from a careful analysis of our \Kwd{C} code.
+This does have the disadvantage that \Kwd{C0}'s memory safety is checked dynamically, which incurs performance overhead.
+Ideally, as described in our 125\% goal, we would like to have a more static verification that memory safety holds, both to provide better performance, and as it provides better results than dynamic checking, since if a dynamic check fails, the program halts, which is less than ideal.
+
+For performance analysis, we plan to evaluate our nameserver for latency and throughput, as these appear to be the most important metrics for \Kwd{DNS} performance.
+We plan to do this using one of two pieces of software\footnote{https://github.com/cobblau/dnsperf, https://github.com/nominum/dnsperf} (confusingly, both named \texttt{dnsperf}), which read queries from a file and then send these requests to a nameserver.
+The latter seems better-suited to our needs, as the former appears to not give latency information in a useful fashion.
+We will compare our results (at various load levels) to existing \Kwd{DNS} software (though we have yet to determine which specific software to take as our baseline).
+One remaining difficulty in evaluation lies in figuring out what sort of query workload to use --- the performance analysis tools we have looked at suggest using an input file with somewhere on the order of one million queries, which is more than can feasibly be generated by hand.
+However, there does not appear to be a standard benchmark for this testing as there are for many other systems.
+We will likely generate large test files in a randomized way, but do not have a good sense of how to make these files accurately reflect realistic usage.
 

### 2017-11-17 proposal.tex
Citation changed for dnsperf.

@@
 For performance analysis, we plan to evaluate our nameserver for latency and throughput, as these appear to be the most important metrics for \Kwd{DNS} performance.
-We plan to do this using one of two pieces of software\footnote{https://github.com/cobblau/dnsperf, https://github.com/nominum/dnsperf} (confusingly, both named \texttt{dnsperf}), which read queries from a file and then send these requests to a nameserver.
+We plan to do this using one of two pieces of software\cite{github:dnsperf1, github:dnsperf2} (confusingly, both named \texttt{dnsperf}), which read queries from a file and then send these requests to a nameserver.
 The latter seems better-suited to our needs, as the former appears to not give latency information in a useful fashion.

### 2017-12-07 src/socket/socket.c
Added main nameserver body.

@@ -106,7 +106,7 @@ c0_int c0_sendto (socket_t sock, c0_array buf, c0_int len,
     // dst_sin_addr, being a 32-bit c0_int, is necessarily the same size as a
     // uint32_t, which dst_addr.sin_addr.s_addr is.
 
-    dst_addr.sin_family = dst_sin_family;
+    dst_addr.sin_family = htons(dst_sin_family);
     dst_addr.sin_port = dst_sin_port;
     dst_addr.sin_addr.s_addr = dst_sin_addr;
     memset(dst_addr.sin_zero, sizeof(dst_addr.sin_zero), 0);

### 2017-12-07 src/socket/socket.c
Minor cleanup.

@@ -99,14 +99,13 @@ c0_int c0_sendto (socket_t sock, c0_array buf, c0_int len,
     assert(SHRT_MIN <= dst_sin_family);
     assert(dst_sin_family <= SHRT_MAX);
 
-    // TODO: USHRT_MIN?
     assert(0 <= dst_sin_port);
     assert(dst_sin_port <= USHRT_MAX);
 
     // dst_sin_addr, being a 32-bit c0_int, is necessarily the same size as a
     // uint32_t, which dst_addr.sin_addr.s_addr is.
 
-    dst_addr.sin_family = htons(dst_sin_family);
+    dst_addr.sin_family = dst_sin_family;
     dst_addr.sin_port = dst_sin_port;
     dst_addr.sin_addr.s_addr = dst_sin_addr;
     memset(dst_addr.sin_zero, sizeof(dst_addr.sin_zero), 0);

### 2017-12-08 paper.tex

--- 2017-11-17 proposal.tex
@@
 \begin{abstract}
-  Please write an abstract.
+  We are implementing a \Kwd{DNS} server in the \Kwd{C0} language,
+  which we intend to evaluate for both safety and performance.
 \end{abstract}
@@
 
-\subsection{What is DNS?}
+\Kwd{Domain Name System} (\Kwd{DNS}) is a hierarchical database that
+maps domain names to IP addresses in a distributed manner augmented
+with caching to achieve a better
+performance~\cite{rfc:1034,rfc:1035,mun-lee:2005}. In fact, there are
+three major parts:
 
-Farzaneh
+\begin{enumerate}
+\item Domain name space with resource records: The domain name space
+  is a tree-like data structure with a root, and nodes labeled with at
+  most 63 characters. The domain name of a label is the sequence of
+  labels on the path from the node to the root. Each node of the tree,
+  associated with a domain name, has a set of resource records.
+
+\item Name servers keep information about a complete sub-tree of the
+  domain name. They may also cache pointers to other name
+  servers.
+
+\item Resolvers are responsible for sending a query to name servers
+  upon clients request and extract information from them. They may
+  also have some parts of the name space cached themselves.  In fact,
+  their role is to provide an interface between applications and DNS
+  in the process of \emph{name resolution}. Name resolution starts
+  when an application program (the host) sends its query for an IP
+  address to a local name server, continues by the local name server
+  which extracts the information in an iterative/recursive manner from
+  different name servers (If it does not already have it in its own
+  cache), and ends when the local name server returns the IP address
+  to the host.
+\end{enumerate}
+
+The messages exchanged between host and name servers adhere to a
+simple format, which is divided into 5 sections: (1) Header, (2)
+Question, (3) Answer, (4) Authority, and (5) Additional.  Each of
+these sections also has its own special format, some of which will be
+covered in more detail in next parts of this document.
 
@@
 
-\subsection{75\%}
+\subsection*{75\%}
 
@@
 
-\subsection{100\%}
+\subsection*{100\%}
 
@@
 
-\subsection{125\%}
+\subsection*{125\%}
 
@@
 
-Our evaluation breaks down into two parts --- one on the safety, and one on performance.
-As safety is a primary goal of this project, we plan to evaluate our code for safety.
-A large part of this will come for free from the memory safety guarantees of \Kwd{C0}, with the rest coming from a careful analysis of our \Kwd{C} code.
-This does have the disadvantage that \Kwd{C0}'s memory safety is checked dynamically, which incurs performance overhead.
-Ideally, as described in our 125\% goal, we would like to have a more static verification that memory safety holds, both to provide better performance, and as it provides better results than dynamic checking, since if a dynamic check fails, the program halts, which is less than ideal.
+Our evaluation breaks down into two parts --- one on the safety, and
+one on performance. As safety is a primary goal of this project, we
+plan to evaluate our code for safety. A large part of this will come
+for free from the memory safety guarantees of \Kwd{C0}, with the rest
+coming from a careful analysis of our \Kwd{C} code. This does have the
+disadvantage that \Kwd{C0}'s memory safety is checked dynamically,
+which incurs performance overhead. Ideally, as described in our 125\%
+goal, we would like to have a more static verification that memory
+safety holds, both to provide better performance, and as it provides
+better results than dynamic checking, since if a dynamic check fails,
+the program halts, which is less than ideal.
 
-For performance analysis, we plan to evaluate our nameserver for latency and throughput, as these appear to be the most important metrics for \Kwd{DNS} performance.
-We plan to do this using one of two pieces of software\cite{github:dnsperf1, github:dnsperf2} (confusingly, both named \texttt{dnsperf}), which read queries from a file and then send these requests to a nameserver.
-The latter seems better-suited to our needs, as the former appears to not give latency information in a useful fashion.
-We will compare our results (at various load levels) to existing \Kwd{DNS} software (though we have yet to determine which specific software to take as our baseline).
-One remaining difficulty in evaluation lies in figuring out what sort of query workload to use --- the performance analysis tools we have looked at suggest using an input file with somewhere on the order of one million queries, which is more than can feasibly be generated by hand.
-However, there does not appear to be a standard benchmark for this testing as there are for many other systems.
-We will likely generate large test files in a randomized way, but do not have a good sense of how to make these files accurately reflect realistic usage.
+For performance analysis, we plan to evaluate our nameserver for
+latency and throughput, as these appear to be the most important
+metrics for \Kwd{DNS} performance. We plan to do this using one of two
+pieces of software\cite{github:dnsperf1, github:dnsperf2}
+(confusingly, both named \texttt{dnsperf}), which read queries from a
+file and then send these requests to a nameserver. The tools mentioned
+in~\cite{github:dnsperf2} seem better-suited to our needs, as those
+of~\cite{github:dnsperf1} appear to not give latency information in a
+useful fashion.
 
-\nocite{rfc:1034, rfc:1035}
+We will compare our results (at various load levels) to existing
+\Kwd{DNS} software (though we have yet to determine which specific
+software to take as our baseline). One remaining difficulty in
+evaluation lies in figuring out what sort of query workload to use ---
+the performance analysis tools we have looked at suggest using an
+input file with somewhere on the order of one million queries, which
+is more than can feasibly be generated by hand. However, there does
+not appear to be a standard benchmark for this testing as there are
+for many other systems. We will likely generate large test files in a
+randomized way, but do not have a good sense of how to make these
+files accurately reflect realistic usage.
+
 \bibliographystyle{abbrv}

### 2017-12-08 paper.tex
start revamping paper

@@
 \usepackage{bytefield}
-\usepackage{fullpage}
 \usepackage{listings}
@@
 \lstset{basicstyle=\ttfamily, columns=flexible}
-
-\usepackage{sectsty}
-\allsectionsfont{\sffamily}
-
-\title{Implementing DNS in C0}
-\author{Farzaneh Derakhshan \and Klaas Pruiksma \and Jonathan Sterling}
-
-\newtheorem{theorem}{Theorem}[section]
-\newtheorem{corollary}{Corollary}[theorem]
-\newtheorem{lemma}[theorem]{Lemma}
 
@@
 
+\usepackage{sectsty}
+\allsectionsfont{\sffamily}
+
+\title{Implementing \Kwd{DNS} in \Kwd{C0}}
+\author{Farzaneh Derakhshan \and Klaas Pruiksma \and Jonathan Sterling}
+
+\newtheorem{theorem}{Theorem}[section]
+\newtheorem{corollary}{Corollary}[theorem]
+\newtheorem{lemma}[theorem]{Lemma}
+
 \begin{document}
@@
 \begin{abstract}
-  We are implementing a \Kwd{DNS} server in the \Kwd{C0} language,
-  which we intend to evaluate for both safety and performance.
+  We have implemented a \Kwd{DNS} server in the \Kwd{C0} language, and
+  evaluated it for both safety and performance.
 \end{abstract}
@@
 
-\subsection{C0 and Concurrent C0}\label{sec:c0}
+\subsection{C0}\label{sec:c0}
 
@@
 there is no need for error-prone manual memory management, since
-\Kwd{C0} is itself garbage-collected. Further safety conditions can be
-enforced with contracts, which are a class of assertions including
-loop invariants, pre- and post-conditions, and dynamically checkable
-correctness conditions for programs.
+\Kwd{C0} is itself garbage-collected.
 
-\Kwd{Concurrent C0} (\Kwd{CC0}) is an extension of \Kwd{C0} that
-offers concurrency based on message passing by providing processes as
-the units of concurrent execution and channels as their message
-buffers~\cite{willsey-prabhu-pfenning:2016}. Concurrent processes
-communicate over channels using a protocol that is disciplined by
-session types~\cite{caires-pfenning-toninho:2013,
-  balzer-pfenning:2017}. In this protocol, each channel is typed such
-that it can either send or receive a message and, thus, it
-deliberately models the client/provider relation.
+Further safety conditions can be enforced with contracts, which are a
+class of assertions including loop invariants, pre- and
+post-conditions, and dynamically checkable correctness conditions for
+programs.
 
-However, the protocol is also able to characterize processes that are
-not completely sequential, and branch into different sequences as they
-encounter choices. This typing discipline also guarantees that data is
-sent and received in the correct order and by the correct processes
-(an invariant called \emph{session fidelity}), and that the
-communication is always one-to-one.
+% \Kwd{Concurrent C0} (\Kwd{CC0}) is an extension of \Kwd{C0} that
+% offers concurrency based on message passing by providing processes as
+% the units of concurrent execution and channels as their message
+% buffers~\cite{willsey-prabhu-pfenning:2016}. Concurrent processes
+% communicate over channels using a protocol that is disciplined by
+% session types~\cite{caires-pfenning-toninho:2013,
+%   balzer-pfenning:2017}. In this protocol, each channel is typed such
+% that it can either send or receive a message and, thus, it
+% deliberately models the client/provider relation.
+
+% However, the protocol is also able to characterize processes that are
+% not completely sequential, and branch into different sequences as they
+% encounter choices. This typing discipline also guarantees that data is
+% sent and received in the correct order and by the correct processes
+% (an invariant called \emph{session fidelity}), and that the
+% communication is always one-to-one.
 
 \subsection{Related Work}
+
+\paragraph{Nail} TODO (Jon) evaluate Nail.
+\nocite{bangert:2014}
 
@@
 
-\section{Goals and Progress}
+% \section{Goals and Progress}
 
-\subsection*{75\%}
+% \subsection*{75\%}
 
-Our 75\% goal is to develop a basic \Kwd{DNS} nameserver in \Kwd{C0}.
+% Our 75\% goal is to develop a basic \Kwd{DNS} nameserver in \Kwd{C0}.
 In order to reduce the scope of this project, we restrict the types of resource data (and by extension, the types of queries) that we handle, and we will only provide support for \Kwd{UDP}-based queries.
@@
 
-The bulk of this goal falls into the two components described in \S\ref{sec:our-project}, along with a small ``main'' component that uses these to actually handle queries.
+% The bulk of this goal falls into the two components described in \S\ref{sec:our-project}, along with a small ``main'' component that uses these to actually handle queries.
 Currently, we have some unmarshalling functions and about a third of the network library implemented.
@@
 
-\subsection*{100\%}
+% \subsection*{100\%}
 
-To extend the above to a 100\% goal, we plan to extend the functionality of our nameserver by adding caching of requests and potentially adding to the types of resource data we support.
+% To extend the above to a 100\% goal, we plan to extend the functionality of our nameserver by adding caching of requests and potentially adding to the types of resource data we support.
 While a nameserver without caching is not particularly useful, we believe that the portion of code most likely to contain memory errors is in some combination of reading data from the network and parsing that data --- in particular, dealing with message compression.
@@
 
-\subsection*{125\%}
+% \subsection*{125\%}
 
-We have two potential 125\% goals, depending on the outcome of our 100\% goal.
+% We have two potential 125\% goals, depending on the outcome of our 100\% goal.
 If our nameserver has good performance relative to its competitors, we will work to provide better justification for the correctness of our nameserver, ideally arriving at a proof that statically verifies that all of our memory accesses are safe.
@@
 
-\section{Evaluation Plan}
+% \section{Evaluation Plan}
 
-Our evaluation breaks down into two parts --- one on the safety, and
-one on performance. As safety is a primary goal of this project, we
-plan to evaluate our code for safety. A large part of this will come
-for free from the memory safety guarantees of \Kwd{C0}, with the rest
-coming from a careful analysis of our \Kwd{C} code. This does have the
-disadvantage that \Kwd{C0}'s memory safety is checked dynamically,
-which incurs performance overhead. Ideally, as described in our 125\%
-goal, we would like to have a more static verification that memory
-safety holds, both to provide better performance, and as it provides
-better results than dynamic checking, since if a dynamic check fails,
-the program halts, which is less than ideal.
+% Our evaluation breaks down into two parts --- one on the safety, and
+% one on performance. As safety is a primary goal of this project, we
+% plan to evaluate our code for safety. A large part of this will come
+% for free from the memory safety guarantees of \Kwd{C0}, with the rest
+% coming from a careful analysis of our \Kwd{C} code. This does have the
+% disadvantage that \Kwd{C0}'s memory safety is checked dynamically,
+% which incurs performance overhead. Ideally, as described in our 125\%
+% goal, we would like to have a more static verification that memory
+% safety holds, both to provide better performance, and as it provides
+% better results than dynamic checking, since if a dynamic check fails,
+% the program halts, which is less than ideal.
 
-For performance analysis, we plan to evaluate our nameserver for
-latency and throughput, as these appear to be the most important
-metrics for \Kwd{DNS} performance. We plan to do this using one of two
-pieces of software\cite{github:dnsperf1, github:dnsperf2}
-(confusingly, both named \texttt{dnsperf}), which read queries from a
-file and then send these requests to a nameserver. The tools mentioned
-in~\cite{github:dnsperf2} seem better-suited to our needs, as those
-of~\cite{github:dnsperf1} appear to not give latency information in a
-useful fashion.
+% For performance analysis, we plan to evaluate our nameserver for
+% latency and throughput, as these appear to be the most important
+% metrics for \Kwd{DNS} performance. We plan to do this using one of two
+% pieces of software\cite{github:dnsperf1, github:dnsperf2}
+% (confusingly, both named \texttt{dnsperf}), which read queries from a
+% file and then send these requests to a nameserver. The tools mentioned
+% in~\cite{github:dnsperf2} seem better-suited to our needs, as those
+% of~\cite{github:dnsperf1} appear to not give latency information in a
+% useful fashion.
 
-We will compare our results (at various load levels) to existing
-\Kwd{DNS} software (though we have yet to determine which specific
-software to take as our baseline). One remaining difficulty in
-evaluation lies in figuring out what sort of query workload to use ---
-the performance analysis tools we have looked at suggest using an
-input file with somewhere on the order of one million queries, which
-is more than can feasibly be generated by hand. However, there does
-not appear to be a standard benchmark for this testing as there are
-for many other systems. We will likely generate large test files in a
-randomized way, but do not have a good sense of how to make these
-files accurately reflect realistic usage.
+% We will compare our results (at various load levels) to existing
+% \Kwd{DNS} software (though we have yet to determine which specific
+% software to take as our baseline). One remaining difficulty in
+% evaluation lies in figuring out what sort of query workload to use ---
+% the performance analysis tools we have looked at suggest using an
+% input file with somewhere on the order of one million queries, which
+% is more than can feasibly be generated by hand. However, there does
+% not appear to be a standard benchmark for this testing as there are
+% for many other systems. We will likely generate large test files in a
+% randomized way, but do not have a good sense of how to make these
+% files accurately reflect realistic usage.
 
@@
 \begin{lstlisting}
-int parse_domain_name(int cursor, int[] data, domain_name_t* dest) {
-  int* len = alloc(int);
+domain_name parse_domain_name(cursor cur, int[] data)
+//@requires cur != NULL;
+//@requires cursor_get(cur) < \length(data) * 32;
+//@requires cursor_get(cur) % 8 == 0;
+{
+    int orig_ix = cursor_get(cur);
+    int len = read_bits(cur, data, 8);
 
-  int orig_cursor = cursor;
-  cursor = read_bits(cursor, data, 8, len);
-
-  if (*len == 0) {
-    // case: done
-    *dest = domain_nil();
-    return cursor;
+    if (len == 0) {
+        return domain_nil();
     }
 
-  if (*len >= 0xc0) {
+    if (len >= 0xc0) {
         // case: pointer
-    int* ptr = alloc(int);
-    // read the remainder of the first 16 bits as a pointer to an offset in octets
-    cursor = read_bits(orig_cursor + 2, data, 14, ptr);
-    // read the last label of the domain name from the pointer:
-    parse_domain_name(*ptr * 8, data, dest);
-    return cursor;
+        // read the last 14 bits of this byte as a
+        // pointer into the message (in octets)
+        cursor_set(cur, orig_ix + 2);
+        int ptr = read_bits(cur, data, 14);
+
+        cursor ptrCursor = cursor_new(ptr * 8);
+        return parse_domain_name(ptrCursor, data);
     }
 
-  // case: ordinary label
-  // now '*len' specifies the number of octets in the label
-  char[] chars = alloc_array(char, *len);
-  for (int i = 0; i < *len; i++) {
-    int* octet = alloc(int);
-    cursor = read_bits(cursor, data, 8, octet);
-    chars[i] = char_chr(*octet);
-  }
-
-  string lbl = string_from_chararray(chars);
-
-  *dest = domain_cons(lbl, *dest);
-  return parse_domain_name(cursor, data, dest);
+    // now we parse an ordinary label; we retreat by one octet
+    // so that read_string can see the length cell
+    cursor_retreat(cur, 8);
+    string lbl = read_string(cur, data);
+    return domain_cons(lbl, parse_domain_name(cur, data));
 }

### 2017-12-08 paper.tex
add a bunch of text to the paper

@@
 \usepackage{amsthm}
-\usepackage[colorlinks=true]{hyperref}
+\usepackage[colorlinks=true, citecolor=red]{hyperref}
 \usepackage{bytefield}
@@
 \usepackage{microtype}
+\usepackage{natbib}
 
@@
 
-\section{Background and Related Work}
+\section{Background}
 
@@
 with caching to achieve a better
-performance~\cite{rfc:1034,rfc:1035,mun-lee:2005}. In fact, there are
+performance~\citep{rfc:1034,rfc:1035,mun-lee:2005}. In fact, there are
 three major parts:
@@
 
-\subsection{C0}\label{sec:c0}
+\subsection*{The \Kwd{C0} Language}\label{sec:c0}
 
@@
 that provides implicit memory safety to
-programmers~\cite{pfenning:c0-reference}. Restrictions on pointer
+programmers~\citep{pfenning:c0-reference}. Restrictions on pointer
 arithmetic and casting, for example, allow the language to detect and
@@
 % the units of concurrent execution and channels as their message
-% buffers~\cite{willsey-prabhu-pfenning:2016}. Concurrent processes
+% buffers~\citep{willsey-prabhu-pfenning:2016}. Concurrent processes
 % communicate over channels using a protocol that is disciplined by
-% session types~\cite{caires-pfenning-toninho:2013,
+% session types~\citep{caires-pfenning-toninho:2013,
 %   balzer-pfenning:2017}. In this protocol, each channel is typed such
@@
 
-\subsection{Related Work}
+\section{Related Work}
 
-\paragraph{Nail} TODO (Jon) evaluate Nail.
-\nocite{bangert:2014}
+\subsection*{Nail: parsing and generating data formats}
 
-\paragraph{The \Kwd{Fox} Project}
+Marshalling and unmarshalling low-level data formats
+is a highly error-prone process, requiring a great degree of care for
+things like byte alignment, sign, etc.; mistakes in this kind of code
+are responsible for numerous security vulnerabilities, especially when
+written in unsafe languages like \Kwd{C}.
+
+To ameliorate this difficulty, \citeauthor{bangert:2014} have
+introduced \Kwd{Nail}, a specialized language for specifying low-level
+data formats. \Kwd{Nail} specifications can then be compiled into real
+parsers and formatters; as part of their evaluation, the authors of
+\Kwd{Nail} have used it to develop a ``authoritative DNS server''.
+
+\Kwd{Nail} has a number of interesting and extremely useful features,
+including \emph{dependent fields}: these are used when the sort of
+data in one field depends on the actual data of another field: for
+instance, when a \verb|length| field is followed by an array of bytes
+of that length.
+%
+Additionally, \Kwd{Nail} supports injecting user-supplied pairs of
+partial inverses to wrap data in the generated parser and formatter;
+this is crucial for compressed or encrypted data.
+
+\paragraph{Evaluation}
+
+We regard the \Kwd{Nail} project as an interesting point in the design
+space, enabling rapid development of tools that interact with
+low-level data formats. As far as safety is concerned, it is necessary
+to trust two things:
+
+\begin{enumerate}
+\item The internal implementation of \Kwd{Nail}, which is written in
+  \Kwd{C}. It is not clear to us what it means to trust \Kwd{C} code,
+  because the ``semantics'' of \Kwd{C} code is completely informal and
+  indefinite.
+\item The user-provided implementations of any functions which are
+  used for compression, decompression, etc. Indeed, there is reason to
+  doubt the correctness of such code, even when written by experts;
+  for instance, the \Kwd{DNS} demo in the \Kwd{Nail} development
+  contains an operator precedence bug~\citep{nigeltao:ticket}.
+\end{enumerate}
+
+The issue described above suggests that while \Kwd{Nail} addresses
+many difficulties, the most error-prone aspects of protocol
+implementation are out of its scope. We think it would be an
+interesting and worthwhile project to pursue a \emph{verified} version
+of \Kwd{Nail} as a library for \Kwd{Coq}~\citep{coq:reference-manual},
+enabling users to verify the correctness of their compression and
+decompression functions relative to a formal approximation of
+\Kwd{C}'s semantics.
+
+Contrary to \Kwd{Nail}, our approach was to code parsers and
+formatters for the \Kwd{DNS} data structures \emph{manually}, but in
+memory-safe and type-safe language. This means that while functional
+correctness is not guaranteed, \emph{safety} is guaranteed.
+
+\subsection*{The \Kwd{Fox} Project}
 An earlier example of an implementation of the DNS protocol in a safe
@@
 undertaken at Carnegie Mellon University in the
-1990s~\cite{biagioni-harper-lee-milnes:1994,
+1990s~\citep{biagioni-harper-lee-milnes:1994,
   biagioni-harper-lee:2001}. Their principal artifact was a library
@@
 
-\paragraph{Other implementations} More recently, safe languages like
-\Kwd{Rust} and \Kwd{Go} have been gaining popularity. Both languages
-have hosted implementations of the \Kwd{DNS}
-protocol~\cite{github:trust-dns,github:miekg-dns}
+\subsection*{Other implementations}
+
+More recently, safe languages
+like \Kwd{Rust} and \Kwd{Go} have been gaining popularity. Both
+languages have hosted implementations of the \Kwd{DNS}
+protocol~\citep{github:trust-dns,github:miekg-dns}
 
@@
 
-typedef struct header* header_t;
+typedef struct header* header;
   \end{lstlisting}
@@
 
-typedef struct domain_name* domain_name_t;
+typedef struct domain_name* domain_name;
 
 struct resource_record {
-  domain_name_t name;
+  domain_name name;
   int type;
@@
 
-typedef struct resource_record* resource_record_t;
+typedef struct resource_record* resource_record;
   \end{lstlisting}
@@
 \begin{lstlisting}
-int parse_domain_name(int cursor, int[] msg, domain_name_t* dest);
-int parse_resource_record(int cursor, int[] msg, resource_record_t dest);
-int parse_header(int cursor, int[] msg, header_t dest);
+domain_name parse_domain_name(cursor cur, int[] data);
+resource_record parse_resource_record(cursor cur, int[] data);
+header parse_header(cursor cur, int[] data);
 \end{lstlisting}
 
-These functions take as an argument a cursor (an index in bits into
-the message), an array of 32-bit words (the entire input message), and
-pointer to a structure into which to write the parsed data; they
-return a new cursor. It is important for the modularity of our
-development that cursors represent not indices into the 32-bit cells
-of the message, but rather bit-indices; this is because specific parts
-of DNS data structures may only be a few bits long, so their parsers
-cannot move the cursor forward a full word's length.
-%
+\paragraph{The cursor abstraction}
+In order to enable a style of coding which is less error-prone than
+manual cursor-passing, we implemented an abstract type of imperative
+cursors, which can be advanced or retreated by a certain number of
+bits. Then, our low-level bit arithmetic utilities can be used to read
+(or write) a certain number of bits at a certain offset from (resp.\
+to) our stream of words.
+
+It is important for the modularity of our development that cursors
+represent not indices into the 32-bit cells of the message, but rather
+bit-indices; this is because specific parts of DNS data structures may
+only be a few bits long, so their parsers cannot move the cursor
+forward a full word's length.
+
 A sample of our parsing code can be found in the supplementary
-appendix to this proposal (\S\ref{appendix:decompression}). Of
-course, we will also implement the inverses to the above functions.
+appendix to this proposal (\S\ref{appendix:decompression}).
+
+\paragraph{The implemented fragment}
+
+We have implemented marshalling and unmarshalling for a sizable
+fragment of the \Kwd{DNS} protocol, omitting only some rare or
+experimental resource record types. Our parser for domain names
+supports compression as specified in~\cite{rfc:1035} (see
+Appendix~\ref{appendix:decompression} for our domain name routine);
+however, during formatting we do not attempt to re-compress domain
+names. This would be necessary for a serious implementation.
+
+\paragraph{Tests}
+
+We have implemented several tests to ensure correct roundtrip behavior
+of the marshalling and unmarshalling routines by hand-coding instances
+of the \Kwd{DNS} message structure.
+
+\section{Evaluation}
+
+\subsection{Correctness and Safety}
+
+By virtue of having written the majority of our \Kwd{DNS} server in
+the memory- and type-safe \Kwd{C0} language, we enjoy certain safety
+guarantees that are either impossible or intractable to achieve for
+mainstream \Kwd{C}-based implementations. In return, we have sacrified
+some performance, but it is hard to tell how much of our performance
+trade-off was due to garbage collection vs.\ the unoptimized character of
+our implementation (but see Section~\ref{sec:performance}).
+
+\paragraph{Contracts in \Kwd{C0}}
+
+One feature of \Kwd{C0} that we have found completely invaluable is
+its notion of \emph{contract}, which are a lightweight form of
+assertion which wraps the input and output boundaries of
+procedures. Consider for instance the following code from our
+low-level bit manipulation library:
+
+\begin{lstlisting}
+int get_mask(int index, int size)
+//@requires size < 32 && size >= 0;
+//@requires index < 32 && index >= 0;
+{
+    return ((1 << size) - 1) << index;
+}
+
+int get_bits(int index, int size, int data)
+//@requires size < 32 && size >= 0;
+//@requires index < 32 && index >= 0;
+{
+    return (data & get_mask(index, size)) >> index;
+}
+
+int set_bits(int data, int index, int size, int value)
+//@requires size < 32 && size >= 0;
+//@requires index < 32 && index >= 0;
+{
+    return (data & ~(get_mask(index, size))) | (value << index);
+}
+\end{lstlisting}
+
+Here, we used contracts to ensure that we would not suffer from bit
+shift range errors. The code did not begin with these contracts, but
+in the course of debugging, these contracts were introduced; in every
+instance, introducing contracts (shifting blame closer and closer to
+call-sites) enabled us to easily determine and liquidate bugs.
+
+The \Kwd{C0} compiler can be instructed to check contracts dynamically
+during execution, or to generate code which ignores them. This is
+helpful, since it is important to have debuggable errors when
+developing, but in production one may not wish to incur the cost of
+dynamic contract checks. While dynamic execution of contract checks
+is already very useful, we feel that more advanced techniques such as
+abstract interpretation and its special case, symbolic execution,
+would be even more useful.
+
+\subsection{Performance}\label{sec:performance}
+
+\textit{Klaas, please write this.}
 
@@
 % metrics for \Kwd{DNS} performance. We plan to do this using one of two
-% pieces of software\cite{github:dnsperf1, github:dnsperf2}
+% pieces of software\citep{github:dnsperf1, github:dnsperf2}
 % (confusingly, both named \texttt{dnsperf}), which read queries from a
@@
 
-\bibliographystyle{abbrv}
+\bibliographystyle{abbrvnat}
 \bibliography{bibtex-references/refs,project}
@@
         // case: pointer
-        // read the last 14 bits of this byte as a
+        // read the last 14 bits of this word as a
         // pointer into the message (in octets)

### 2017-12-10 src/socket/socket.c
Added network byte order to socket interface.

@@ -12,7 +12,11 @@
 #include <sys/socket.h>
 #include <sys/types.h>
 
+#define INT16_T_MAX (0x7FFF)
+#define INT16_T_MIN (-(0x8000))
+
 #define INT32_T_MAX 0x7FFFFFFF
+#define INT32_T_MIN 0x80000000
 
 #define ROUND_TO_4(x) ((((x) + 3) / 4) * 4)
 
@@ -192,5 +196,41 @@ c0_int poll_write (socket_t sock, c0_int timeout) {
     }
     
     assert(0 == rv || (1 == rv && sockfd.revents & POLLOUT));
+    return rv;
+}
+
+c0_int c0_ntohl(c0_int netlong) {
+    assert (INT32_T_MIN <= netlong && netlong <= INT32_T_MAX)
+
+    c0_int rv = ntohl(netlong);
+    
+    assert (INT32_T_MIN <= rv && rv <= INT32_T_MAX)
+    return rv;
+}
+
+c0_int c0_ntohs(c0_int netshort) {
+    assert (INT16_T_MIN <= netlong && netlong <= INT16_T_MAX)
+
+    c0_int rv = ntohl(netlong);
+    
+    assert (INT16_T_MIN <= rv && rv <= INT16_T_MAX)
+    return rv;
+}
+
+c0_int c0_htonl(c0_int netlong) {
+    assert (INT32_T_MIN <= netlong && netlong <= INT32_T_MAX)
+
+    c0_int rv = htonl(netlong);
+    
+    assert (INT32_T_MIN <= rv && rv <= INT32_T_MAX)
+    return rv;
+}
+
+c0_int c0_htons(c0_int netshort) {
+    assert (INT16_T_MIN <= netlong && netlong <= INT16_T_MAX)
+
+    c0_int rv = htonl(netlong);
+    
+    assert (INT16_T_MIN <= rv && rv <= INT16_T_MAX)
     return rv;
 }

### 2017-12-10 src/socket/socket.c
network byte order functions updated.

@@ -7,6 +7,7 @@
 
 #include <c0runtime.h>
 
+#include <arpa/inet.h>
 #include <netinet/in.h>
 
 #include <sys/socket.h>
@@ -200,37 +201,37 @@ c0_int poll_write (socket_t sock, c0_int timeout) {
 }
 
 c0_int c0_ntohl(c0_int netlong) {
-    assert (INT32_T_MIN <= netlong && netlong <= INT32_T_MAX)
+    assert (INT32_T_MIN <= netlong && netlong <= INT32_T_MAX);
 
     c0_int rv = ntohl(netlong);
     
-    assert (INT32_T_MIN <= rv && rv <= INT32_T_MAX)
+    assert (INT32_T_MIN <= rv && rv <= INT32_T_MAX);
     return rv;
 }
 
 c0_int c0_ntohs(c0_int netshort) {
-    assert (INT16_T_MIN <= netlong && netlong <= INT16_T_MAX)
+    assert (INT16_T_MIN <= netshort && netshort <= INT16_T_MAX);
 
-    c0_int rv = ntohl(netlong);
+    c0_int rv = ntohs(netshort);
     
-    assert (INT16_T_MIN <= rv && rv <= INT16_T_MAX)
+    assert (INT16_T_MIN <= rv && rv <= INT16_T_MAX);
     return rv;
 }
 
-c0_int c0_htonl(c0_int netlong) {
-    assert (INT32_T_MIN <= netlong && netlong <= INT32_T_MAX)
+c0_int c0_htonl(c0_int hostlong) {
+    assert (INT32_T_MIN <= hostlong && hostlong <= INT32_T_MAX);
 
-    c0_int rv = htonl(netlong);
+    c0_int rv = htonl(hostlong);
     
-    assert (INT32_T_MIN <= rv && rv <= INT32_T_MAX)
+    assert (INT32_T_MIN <= rv && rv <= INT32_T_MAX);
     return rv;
 }
 
-c0_int c0_htons(c0_int netshort) {
-    assert (INT16_T_MIN <= netlong && netlong <= INT16_T_MAX)
+c0_int c0_htons(c0_int hostshort) {
+    assert (INT16_T_MIN <= hostshort && hostshort <= INT16_T_MAX);
 
-    c0_int rv = htonl(netlong);
+    c0_int rv = htons(hostshort);
     
-    assert (INT16_T_MIN <= rv && rv <= INT16_T_MAX)
[…]

### 2017-12-10 src/socket/socket.c
Added network byte order for metadata.

@@ -55,7 +55,7 @@ c0_int c0_recvfrom (socket_t sock, c0_array buf, c0_int len,
                     c0_pointer src_sin_port,
                     c0_pointer src_sin_addr) {
     struct sockaddr_in src_addr;
-    socklen_t addrlen;
+    socklen_t addrlen = sizeof(struct sockaddr_in);
     size_t read_len = (size_t) len;
     int flags = MSG_TRUNC;
 
@@ -85,9 +85,9 @@ c0_int c0_recvfrom (socket_t sock, c0_array buf, c0_int len,
     /* TODO */
     free(tmp_buf);
 
-    *((c0_int*) c0_deref(src_sin_family)) = src_addr.sin_family;
-    *((c0_int*) c0_deref(src_sin_port)) = src_addr.sin_port;
-    *((c0_int*) c0_deref(src_sin_addr)) = src_addr.sin_addr.s_addr;
+    *((c0_int*) c0_deref(src_sin_family)) = ntohs(src_addr.sin_family);
+    *((c0_int*) c0_deref(src_sin_port)) = ntohs(src_addr.sin_port);
+    *((c0_int*) c0_deref(src_sin_addr)) = ntohl(src_addr.sin_addr.s_addr);
 
     return rv;
 }
@@ -110,9 +110,9 @@ c0_int c0_sendto (socket_t sock, c0_array buf, c0_int len,
     // dst_sin_addr, being a 32-bit c0_int, is necessarily the same size as a
     // uint32_t, which dst_addr.sin_addr.s_addr is.
 
-    dst_addr.sin_family = dst_sin_family;
-    dst_addr.sin_port = dst_sin_port;
-    dst_addr.sin_addr.s_addr = dst_sin_addr;
+    dst_addr.sin_family = htons(dst_sin_family);
+    dst_addr.sin_port = htons(dst_sin_port);
+    dst_addr.sin_addr.s_addr = htonl(dst_sin_addr);
     memset(dst_addr.sin_zero, sizeof(dst_addr.sin_zero), 0);
 
     /* TODO: Find a way to not allocate here. */

### 2017-12-11 paper.tex
fix typo

@@
 Contrary to \Kwd{Nail}, our approach was to code parsers and
-formatters for the \Kwd{DNS} data structures \emph{manually}, but in
+formatters for the \Kwd{DNS} data structures \emph{manually}, but in a
 memory-safe and type-safe language. This means that while functional

### 2017-12-12 src/socket/socket.c
Bugs in byte order functions fixed. Tests now run properly.

@@ -5,6 +5,8 @@
 #include <string.h>
 #include <unistd.h>
 
+#include <stdio.h>
+
 #include <c0runtime.h>
 
 #include <arpa/inet.h>
@@ -13,11 +15,11 @@
 #include <sys/socket.h>
 #include <sys/types.h>
 
-#define INT16_T_MAX (0x7FFF)
-#define INT16_T_MIN (-(0x8000))
+#define INT16_T_MAX ((int32_t)32767)
+#define INT16_T_MIN ((int32_t)-32767)
 
-#define INT32_T_MAX 0x7FFFFFFF
-#define INT32_T_MIN 0x80000000
+#define INT32_T_MAX ((int32_t)0x7FFFFFFF)
+#define INT32_T_MIN ((int32_t)0x80000000)
 
 #define ROUND_TO_4(x) ((((x) + 3) / 4) * 4)
 
@@ -201,7 +203,7 @@ c0_int poll_write (socket_t sock, c0_int timeout) {
 }
 
 c0_int c0_ntohl(c0_int netlong) {
-    assert (INT32_T_MIN <= netlong && netlong <= INT32_T_MAX);
+    assert (0 <= (uint32_t) netlong && (uint32_t) netlong <= 0xFFFFFFFF);
 
     c0_int rv = ntohl(netlong);
     
@@ -210,7 +212,7 @@ c0_int c0_ntohl(c0_int netlong) {
 }
 
 c0_int c0_ntohs(c0_int netshort) {
-    assert (INT16_T_MIN <= netshort && netshort <= INT16_T_MAX);
+    assert (0 <= (uint16_t) netshort && (uint16_t) netshort <= 0xFFFF);
 
     c0_int rv = ntohs(netshort);
     
@@ -223,7 +225,7 @@ c0_int c0_htonl(c0_int hostlong) {
 
     c0_int rv = htonl(hostlong);
     
-    assert (INT32_T_MIN <= rv && rv <= INT32_T_MAX);
+    assert (0 <= (uint32_t) rv && (uint32_t) rv <= 0xFFFFFFFF);
     return rv;
 }
 
@@ -231,7 +233,7 @@ c0_int c0_htons(c0_int hostshort) {
     assert (INT16_T_MIN <= hostshort && hostshort <= INT16_T_MAX);
 
     c0_int rv = htons(hostshort);
-    
-    assert (INT16_T_MIN <= rv && rv <= INT16_T_MAX);
+
[…]

### 2017-12-12 src/socket/socket.c
We appear to have a working DNS server.

@@ -1,4 +1,5 @@
 #include <assert.h>
+#include <errno.h>
 #include <poll.h>
 #include <stdlib.h>
 #include <limits.h>
@@ -32,20 +33,35 @@ socket_t dgram_socket() {
     if (sock < 0) {
         c0_abort("Socket creation failed.");
     }
-
     return sock;
+}
+
+c0_int c0_bind(socket_t sock, c0_int port) {
+    struct sockaddr_in addr;
+    addr.sin_family = AF_INET;
+    addr.sin_port = htons(port);
+    addr.sin_addr.s_addr = htonl(INADDR_ANY);
+    int rv = bind(sock, (struct sockaddr *) &addr, sizeof(struct sockaddr_in));
+
+    if (rv < 0) {
+        c0_abort("Socket bind failed.");
+    }
+
+    return rv;
 }
 
 c0_int recv_msglen (socket_t sock) {
     struct sockaddr src_addr;
     socklen_t addrlen;
-    size_t len = 0;
+    size_t len = 512;
+    char buf[512];
     int flags = MSG_TRUNC | MSG_PEEK;
 
-    ssize_t pkt_size = recvfrom(sock, NULL, len, flags, &src_addr, &addrlen);
-    if (-1 == pkt_size) {
+    ssize_t pkt_size = recvfrom(sock, buf, len, flags, &src_addr, &addrlen);
+    if (pkt_size < 0) {
         c0_error("recv_msglen failed.");
     }
+    assert(0 <= pkt_size);
     assert(pkt_size <= INT32_T_MAX);
 
     c0_int rv = (c0_int) pkt_size;
@@ -216,7 +232,7 @@ c0_int c0_ntohs(c0_int netshort) {
 
     c0_int rv = ntohs(netshort);
     
-    assert (INT16_T_MIN <= rv && rv <= INT16_T_MAX);
+    assert (0 <= (uint16_t) rv && (uint16_t) rv <= 0xFFFF);
     return rv;
 }
 
@@ -230,10 +246,10 @@ c0_int c0_htonl(c0_int hostlong) {
 }
 
 c0_int c0_htons(c0_int hostshort) {
-    assert (INT16_T_MIN <= hostshort && hostshort <= INT16_T_MAX);
+    assert (0 <= (uint16_t) hostshort && (uint16_t) hostshort <= 0xFFFF);
 
     c0_int rv = htons(hostshort);
 
[…]

### 2017-12-12 Presentation.tex
Slides

\documentclass[slidestop]{beamer}

\usepackage{beamerthemesplit}
\usepackage{mathpartir}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage{grffile}
\usepackage{graphics}
\usepackage{pstricks}
\usepackage{bytefield}
\usepackage{listings}

\lstset{language=C}
\lstset{basicstyle=\ttfamily, columns=flexible}

\newcommand\Kwd[1]{{\sffamily\bfseries{#1}}}
\newcommand\EOf[4]{{#1}\vdash_{#2}{#3}:{#4}}
\newcommand\MOf[4]{{#1}\vdash_{#2}{#3}\div{#4}}
\newcommand\IsProc[3]{{#1}\vdash_{#2}{#3}\ \mathsf{proc}}
\newcommand\IsAction[3]{{#1}\vdash_{#2}{#3}\ \mathsf{action}}

\title{Implementing DNS in C0}
\author{Jonathan Sterling \and Klaas Pruiksma \and Farzaneh Derakhshan}

\begin{document}

\frame{
   \begin{center}
    \huge{Implementing DNS in C0}\\
    \vspace{48pt}
    \Large{15-712 Project}\\
    \vspace{24pt}
    \large{\today}
  \end{center}
}

\frame{\frametitle{Background-DNS}
 \begin{center}

  {\bf \red Map domain names to IP addresses in a distributed manner. Consists of: }
 \end{center} 
\begin{itemize}
\item Domain name space with reource records, a tree-like data structure with a root, and nodes labeled with at most 63 characters.
\item Name servers keep information about a complete sub-tree of the domain name. They may also cache pointers to other name servers.
\item Resolvers that provide an interface between applications and DNS in the process of name resolution
\end{itemize}
}
\frame{\frametitle{DNS-Name Resolution}
\begin{itemize}
\item A user program sends a name query to the resolver of the DNS client.

\item The DNS resolver looks up the local domain name cache for a match. If a match is found, it sends the corresponding IP address back. If not, it sends a query to the DNS server.

\item The DNS server looks up the corresponding IP address of the domain name in its DNS database. If no match is found, it sends a query to a higher level DNS server. This process continues until a result, whether successful or not, is returned.

\item The DNS client returns the resolution result to the application after receiving a response from the DNS server.
 \end{itemize}
}
\frame{\frametitle{Background-C0}
 \begin{center}

  {\bf \red C0 programming language is a restricted subset of C, that provides:}
\begin{itemize}
\item Implicit {\bf memory safety}.\\
\item Restrictions on {\bf pointer arithmetic} and {\bf casting}: 
\begin{itemize}
\item prevent any attempt to access an array out of its
bounds.\\
\end{itemize}
\item {\bf Garbage-collection} 
\begin{itemize} 
\item no need for error-prone manual memory management.\\
\end{itemize}
\item {\bf Contracts}: dynamically checks correctness conditions for programs\\
\end{itemize}
 \end{center}
}

\frame{\frametitle{Related Work-Nail}
{\bf \red Nail}: A specialized language for specifying low-level data formats, have been used to develop a {\bf \red “authoritative DNS server”.} \vspace{12pt}

 The most error-prone aspects are out of Nail's scope:\\
\begin{itemize}
\item {\bf Internal implementation of Nail} written in {\bf C}. 
\begin{itemize}
\item “Semantics” of C code is informal and indefinite.
\end{itemize}
\item {\bf User-provided implementation} of compression and decompression functions. 
\begin{itemize}
\item An operator precedence bug in the Nail's DNS demo .
\end{itemize}
\end{itemize}
\vspace{12pt}

Conntrary to Nail, we code parsers and formatters {\bf manually}, but in a {\bf memory-safe} and {\bf type-safe language}.

\begin{center} 
{\bf \red While functional correctness is not guaranteed, safety is guaranteed.}
 \end{center}
}
\frame{\frametitle{Related Work-Fox}
{\bf \red Fox }: Implementation of the DNS protocol in a safe and formally specified language,
\begin{itemize}
\item undertaken at Carnegie Mellon University in the 1990s.\\
\item The FoXNet library implemented

\begin{center} An array of low-level network infrastructure (TCP, IP, ARP and DNS) in the Standard ML language.\end{center}
\item Unlike our development, FoxNet had a highly modular character.
\end{itemize}
}

\frame{\frametitle{Our Project}
 \begin{center}
  {\bf \red Developing an implementation of the DNS protocol in the memory-safe C0 language}
 \end{center}
Two principal components:
\begin{itemize}
\item Network Interface
\item Marshalling and Unmarshalling
\end{itemize}

}

\frame{\frametitle{Network Interface}
C0 has a very limited standard library.
 \begin{itemize}
   \item We write a networking library for C0.\vspace{12pt}
   \item Due to limitations of the language, This library cannot be written entirely in C0.\vspace{12pt}
\item We write an interface between C0 code and the C sockets library with a collection of small functions in C.\vspace{12pt}
\item Add on top of that, the C0 portion of the library, to provide a higher-level interface for network communication.\vspace{12pt}
  \end{itemize}
\vspace{12pt}}
\frame{\frametitle{Network Interface-Cont'd}
\begin{center}
{\bf \red ensure that any memory safety violations occur outside of our project}
\end{center}
\begin{itemize}
\item  Keep the C portion limited in scope and size:
\begin{itemize}
\item  By checking that there are no calls to malloc or similar functions.
\end{itemize}
\item Any accesses to C0-allocated memory are done so safely.(less easily checked)
\begin{itemize}
\item By using a collection of functions for safe memory access provided to libraries by the C0 runtime.
\end{itemize}
\end{itemize}
}
\frame{\frametitle{Network Interface-Cont'd}  
\begin{center}
limited in the scope to DNS, so we
\end{center}
\begin{itemize}
\item Focus on the functions for datagram-based sockets,
\item Do not expose the full functionality of the sockets library, 
\item A limited interface to the poll function of the C standard library in the same way to avoid busy-waiting.\\
\end{itemize}
\begin{center}
Sufficient to implement our nameserver\\
Additional library functionality can be added using these methods.
\end{center}
}

\frame{\frametitle{Marshalling and Unmarshalling}

 \begin{itemize}
   \item C0 has only a single type (32-bit) integers.\vspace{12pt}
   \item The requests and responses to the network interface formatted as arrays of 32-bit integers.\vspace{12pt}
\item The marshalling-and-unmarshalling module unpack this data into structured representations of core DNS datatypes.\vspace{24pt}
  \end{itemize}}
\frame{\frametitle{Marshalling and Unmarshalling-cont'd}
This figure shows the bit-level layout of DNS message headers:
\begin{figure}
  \centering
  \begin{bytefield}[bitwidth=1.5em]{16}
    \wordbox{1}{\texttt{ID}}
    \\
    \bitbox{1}{\texttt{QR}}
    \bitbox{4}{\texttt{Opcode}}
    \bitbox{1}{\texttt{AA}}
    \bitbox{1}{\texttt{TC}}
    \bitbox{1}{\texttt{RD}}
    \bitbox{1}{\texttt{RA}}
    \bitbox{3}{\texttt{Z}}
    \bitbox{4}{\texttt{RCODE}}
    \\
    \wordbox{1}{\texttt{QDCOUNT}}
    \\
    \wordbox{1}{\texttt{ANCOUNT}}
    \\
    \wordbox{1}{\texttt{NSCOUNT}}
    \\
    \wordbox{1}{\texttt{ARCOUNT}}
  \end{bytefield}
  \caption{The layout specification of message headers in
    DNS.}\label{fig:layout-message-header}
\end{figure}}

\begin{frame}[fragile]\frametitle{Marshalling and Unmarshalling}
 How we code the header structure in our $C0$ implementation:{\tiny
\begin{figure}
  \begin{lstlisting}[frame=single]
struct header {
  int id;
  int qr;
  int opcode;
  int aa;
  int tc;
  int rd;
  int ra;
  int z;
  int rcode;
  int qdcount;
  int ancount;
  int nscount;
  int arcount;
};
typedef struct header* header;
  \end{lstlisting}

\caption{Example DNS data structures in our C0 implementation.}\label{fig:c0-data-structures}
\end{figure}}
\end{frame}
\begin {frame}[fragile]\frametitle{Marshalling and Unmarshalling}
\begin{itemize}
\item To convert between these two we implement functions of the following kind:
\end{itemize}
{\small
\begin{lstlisting}
domain_name parse_domain_name(cursor cur, int[] data);
resource_record parse_resource_record(cursor cur, int[] data);
header parse_header(cursor cur, int[] data);
\end{lstlisting}}

\end{frame}
\frame{\frametitle{Marshalling and Unmarshalling-Cont'd}
\begin{itemize}
   \item To enable a less error-prone style of coding than manual cursor-passing, we implemented {\bf an abstract type of imperative
cursors}.
\item A cursor can be advanced or retreated by a certain number of bits. \\
\item Then, our low-level bit arithmetic utilities can be used to read (or write) a certain number of bits at a certain offset from (resp. to) our stream of words.\\
\item cursors are bit-indices rather than indices into the 32-bit cells of the message.
\item Because specific parts of DNS data structures may only be a few bits long, so their parsers cannot move the cursor forward a full word’s length.
  \end{itemize}
We have implemented marshalling and unmarshalling for a sizable fragment of the DNS protocol, omitting only some rare or experimental resource record types.\\
Our parser for domain names supports compression\\
During formatting we do not attempt to re-compress domain names. 
}

\frame{\frametitle{Evaluation-correctness and safety}

\begin{center}
we enjoy certain safety guarantees that are either impossible or intractable to achieve for mainstream C-based implementations.
\end{center}

 \begin{itemize}
 \item writing the majority of our DNS server in the memory- and type- safe C0 language, 
\item In return, we have sacrified some performance. 
\begin{itemize}
\item Hard to tell how much of our performance trade-off was due to garbage collection vs. the unoptimized character of our implementation
\end{itemize}
  \end{itemize}
\vspace{12pt}}
\begin{frame}[fragile]\frametitle{Evaluation-correctness and safety-Contracts}
 invaluable notion of {\bf contract}:\\
Here, we used contracts to ensure that we would not suffer from bit shift range
errors. 
a lightweight form of assertion which wraps the input and output boundaries of procedures.\\
{\small \begin{lstlisting}
int get_mask(int index, int size)
//@requires size < 32 && size >= 0;
//@requires index < 32 && index >= 0;
{
    return ((1 << size) - 1) << index;
}
int get_bits(int index, int size, int data)
//@requires size < 32 && size >= 0;
//@requires index < 32 && index >= 0;
{
    return (data & get_mask(index, size)) >> index;
}

\end{lstlisting}}

\end{frame}

\frame{\frametitle{Evaluation-correctness}

  The code did not begin with these contracts, but in the course of debugging,
these contracts were introduced; in every instance, introducing contracts (shifting
blame closer and closer to call-sites) enabled us to easily determine and liquidate bugs.\vspace{12pt}
   \begin{itemize}
   \item The C0 compiler can be instructed to check contracts dynamically during execution,
or to generate code which ignores them. This is helpful, since it is important to have
debuggable errors when developing, but in production one may not wish to incur the
cost of dynamic contract checks. While dynamic execution of contract checks is already
very useful, we feel that more advanced techniques such as abstract interpretation and
its special case, symbolic execution, would be even more useful.\vspace{12pt}
  \end{itemize}
}
\frame{\frametitle{Evaluation-Performance}

}

\end{document}

### 2017-12-13 paper.tex
Updated networking section.

@@
 
-In its current state, \Kwd{C0} has a very limited standard library.
-In particular, while there are libraries for input and output to files and to console, there is no networking library.
-It may be possible to implement networking using file I/O, but we do not believe that this is reasonable.
-
-As such, we will write a networking library for \Kwd{C0}.
-Due to limitations of the language, however, this library cannot be written entirely in \Kwd{C0} itself.
-Instead, we will write a collection of small functions in \Kwd{C} which act as an interface between \Kwd{C0} code and the \Kwd{C} sockets library.
-On top of this, we can then write the \Kwd{C0} portion of the library, which will provide a higher-level interface for network communication than that provided by the \Kwd{C} portion of the library.
+We have implemented a (limited) networking library for \Kwd{C0}.
+Due to limitations of the language, however, this library is not written entirely in \Kwd{C0} itself.
+Instead, we wrote a collection of small functions in \Kwd{C} which act as an interface between \Kwd{C0} code and the \Kwd{C} sockets library.
 
 By splitting the library into these two parts, we can keep the \Kwd{C} portion of our codebase limited in scope.
-In particular, we plan to ensure that we never dynamically allocate memory from \Kwd{C} and that any accesses to \Kwd{C0}-allocated memory are done so safely.
-The former is easily checked for, simply by checking that there are no calls to malloc or similar functions, while the latter can be achieved (but less easily checked) by using a collection of functions for safe memory access provided to libraries by the \Kwd{C0} runtime.
+While our original plan was to never dynamically allocate memory in this code, we found that it was infeasible to use the socket functions without either dynamically allocating memory or violating other portions of our safety guarantees.
+As such, dynamically allocating memory was the safer choice, and so we have one allocation in each of the \texttt{sendto} and \texttt{recvfrom} functions.
+These are the only potentially unsafe regions of memory used in the application, and so we closely hand-checked their use for correctness.
+Additionally, we were able to ensure that any accesses to \Kwd{C0}-allocated memory are done so  using a collection of functions for safe memory access provided to libraries by the \Kwd{C0} runtime, ensuring safety of these accesses.
 As a result of this and the small size of the \Kwd{C} portion of the codebase, we can easily check all of the \Kwd{C} code for memory safety.
@@
 
-Since this project is limited in scope to \Kwd{DNS}, we will not expose the full functionality of the sockets library in our network interface.
-We will instead focus on the functions that create and work with datagram-based sockets, as most of the network communication needed for \Kwd{DNS} is done via the datagram protocol \Kwd{UDP}.
-Additionally, to avoid busy-waiting, we will provide a limited interface to the \texttt{poll} function of the \Kwd{C} standard library in much the same way.
-We believe that this is sufficient to implement our nameserver, but additional library functionality can be added using these methods if necessary.
+Since this project is limited in scope to \Kwd{DNS}, we do not expose the full functionality of the sockets library in our network interface.
+We implemented functions that create and work with datagram-based sockets, as most of the network communication needed for \Kwd{DNS} is done via the datagram protocol \Kwd{UDP}.
+Additionally, to avoid busy-waiting, we provide a limited interface to the \texttt{poll} function of the \Kwd{C} standard library in much the same way.
+Finally, we add \Kwd{C0} interfaces for the collection of functions that translate between network and host byte order, in order to shift as much work as possible into \Kwd{C0}, rather than changing byte order in our \Kwd{C} code.
 
@@
 guarantees that are either impossible or intractable to achieve for
-mainstream \Kwd{C}-based implementations. In return, we have sacrified
+mainstream \Kwd{C}-based implementations. In return, we have sacrificed
 some performance, but it is hard to tell how much of our performance

### 2017-12-13 paper.tex
Images added to paper.
Updated paper. Should be in a complete-enough state now.

@@
 \usepackage{amsthm}
+\usepackage{graphicx}
 \usepackage[colorlinks=true, citecolor=red]{hyperref}
@@
 
-\textit{Klaas, please write this.}
+In order to evaluate the performance of our nameserver (\Kwd{c0ns}), we used \Kwd{resperf}, from the freely available set of DNS benchmarking tools provided by~\cite{github:dnsperf2}.
+Since our nameserver is recursive, rather than authoritative (that is, it relies on remote nameservers for the ground truth), \Kwd{resperf} was a more suitable performance tool than \Kwd{dnsperf}.
+
+As we were unable to find any existing data sets with realistic loads for testing, we created a data set of approximately 700,000 entries of A queries (which appear to be the most common) by turning each word in the system dictionary into three URLs to query.
+We believe that this meets the criteria suggested for performance testing in that it has a good mixture of real and nonexistent domain names, and as we need to be able to read resource records other than A when receiving messages from the remote servers that we query.
+Ideally, we would like to use a more varied dataset that reflects real-world usage, but this appears to work well enough for performance testing.
+One other issue with our dataset is that all queries in it are unique.
+While this does not matter for our nameserver, as it does not cache, were we to add caching, we would like to use a dataset with repetitions in order to test the effect of caching on performance.
+
+We began by testing with the default settings of \Kwd{resperf}, wherein the number of queries being sent per second increases linearly from 0 to 100,000 over the course of a minute.
+We found, however, that this did not yield useful results, as the throughput capacity of \Kwd{c0ns} is so low by comparison.
+After adjusting parameters to get useful data, we found that setting a maximum of 100 queries per second over the same time interval of 60 seconds was suitable.
+We then tested both the checked (safe) and unchecked (not safe) versions of \Kwd{c0ns} with these parameters.
+Interestingly, the difference between the performance of the checked version (shown in Figure~\cite{fig:checked-perf}) and the performance of the unchecked version (shown in Figure~\cite{fig:unchecked-perf}) was minimal, suggesting that contract-checking provides little to no overhead at these speeds (though it is likely that with higher base throughput, the difference would become more evident).
+
+We also note that our latency remains low up until about 20 seconds into the test, at which point the number of queries per second is around 33.
+The latency graphs thus support the conclusion that we can maintain sustained loads of around 30 queries per second (as suggested as well by the throughput graph), but past that point, we appear to have saturated the server, as latency continues to increase up to 45 seconds (at which point \Kwd{resperf} considers queries to have failed).
+
+Because of the low throughput of our system, we did not find it useful to test against commercial nameservers, relying instead on previous work such as \cite{fagin2017making}, which suggests that commercial nameservers can handle loads on the order of 20,000 to 80,000 queries per second.
+With such a large magnitude of difference, even if commercial nameservers perform substantially worse on our benchmark than on the benchmarks that others have used, we still have around 1,000-3,000 times less throughput.
+
+\begin{figure}[h]
+    \includegraphics[width=0.5\textwidth]{unchecked_rate.png}
+    \includegraphics[width=0.5\textwidth]{unchecked_latency.png}
+    \caption{Performance of the unchecked version of \Kwd{c0ns}}\label{fig:unchecked-perf}
+\end{figure}
+
+\begin{figure}[h]
+    \includegraphics[width=0.5\textwidth]{checked_rate.png}
+    \includegraphics[width=0.5\textwidth]{checked_latency.png}
+    \caption{Performance of the checked version of \Kwd{c0ns}}\label{fig:checked-perf}
+\end{figure}
+
+\section{Goals and Attainment}
+
+We believe that this project meets our 75\% goal very solidly, but falls short of the 100\% goal.
+The primary obstacle to finishing the 100\% goal was that we underestimated the amount of time it would take to clear up bugs relating to network byte order, which were the source of much debugging time.
+While we believe that adding caching would not be particularly difficult, given the structure of our nameserver, there simply was not enough time to do so.
+
+Moreover, from our testing, it appears that there are more pressing issues to fix with \Kwd{c0ns}.
+In particular, we ran into some issues while testing in that while we implement all of the resource record types that we expect to get from a client, sometimes we get unexpected resource record types from the remote nameserver that we query.
+In particular, we do not handle NS resource records, which are used to inform recursive nameservers of which server is authoritative for the query that was made if the remote server is not authoritative.
+Adding support for these seems to be a much higher priority than adding caching.
+Additionally, our low performance even on benchmarks that do not benefit at all from caching suggests that perhaps we should focus on performance on unique queries before adding support for caching.
+In particular, we believe that performance could be greatly improved by adding support for concurrently processing requests, since resolving a query is network-bound (while we wait for the remote server to get back to us), and so should not block CPU-bound tasks like parsing other incoming queries.
 
@@
 
+\clearpage
+
 \bibliographystyle{abbrvnat}

### 2017-12-13 paper.tex
fix figure refs

@@
 We then tested both the checked (safe) and unchecked (not safe) versions of \Kwd{c0ns} with these parameters.
-Interestingly, the difference between the performance of the checked version (shown in Figure~\cite{fig:checked-perf}) and the performance of the unchecked version (shown in Figure~\cite{fig:unchecked-perf}) was minimal, suggesting that contract-checking provides little to no overhead at these speeds (though it is likely that with higher base throughput, the difference would become more evident).
+Interestingly, the difference between the performance of the checked version (shown in Figure~\ref{fig:checked-perf}) and the performance of the unchecked version (shown in Figure~\ref{fig:unchecked-perf}) was minimal, suggesting that contract-checking provides little to no overhead at these speeds (though it is likely that with higher base throughput, the difference would become more evident).
 

### 2017-12-13 paper.tex
Removed section on NS records from paper.

@@
 Moreover, from our testing, it appears that there are more pressing issues to fix with \Kwd{c0ns}.
-In particular, we ran into some issues while testing in that while we implement all of the resource record types that we expect to get from a client, sometimes we get unexpected resource record types from the remote nameserver that we query.
-In particular, we do not handle NS resource records, which are used to inform recursive nameservers of which server is authoritative for the query that was made if the remote server is not authoritative.
-Adding support for these seems to be a much higher priority than adding caching.
-Additionally, our low performance even on benchmarks that do not benefit at all from caching suggests that perhaps we should focus on performance on unique queries before adding support for caching.
+In  particular, our low performance even on benchmarks which do not benefit at all from caching suggests that perhaps we should focus on performance on unique queries before adding support for caching.
 In particular, we believe that performance could be greatly improved by adding support for concurrently processing requests, since resolving a query is network-bound (while we wait for the remote server to get back to us), and so should not block CPU-bound tasks like parsing other incoming queries.

### 2017-12-13 paper.tex
add further remarks about 75%/100% goals

@@
 
-We believe that this project meets our 75\% goal very solidly, but falls short of the 100\% goal.
-The primary obstacle to finishing the 100\% goal was that we underestimated the amount of time it would take to clear up bugs relating to network byte order, which were the source of much debugging time.
-While we believe that adding caching would not be particularly difficult, given the structure of our nameserver, there simply was not enough time to do so.
+We believe that this project meets our 75\% goal very solidly, but
+falls short of the 100\% goal.\footnote{Our 75\% goal was to simply implement
+  a working \Kwd{DNS} server; our 100\% goal was to extend this basic
+  server with support for caching, and for further resource record
+  types; we would also have liked to support compressing domain
+  names. In our implementation, we supported only \emph{decompressing}
+  domain names, which is allowed by the \Kwd{DNS} specification, but
+  is not optimal~\citep{rfc:1035}.} The primary obstacle to finishing
+the 100\% goal was that we underestimated the amount of time it would
+take to clear up bugs relating to network byte order, which were the
+source of much debugging time. While we believe that adding caching
+would not be particularly difficult, given the structure of our
+nameserver, there simply was not enough time to do so.
 

### 2017-12-13 paper.tex
update example code in paper

@@
 
-    if (len >= 0xc0) {
+    if ((len & 0xc0) == 0xc0) {
         // case: pointer
-        // read the last 14 bits of this word as a
-        // pointer into the message (in octets)
-        cursor_set(cur, orig_ix + 2);
-        int ptr = read_bits(cur, data, 14);
+        // read the remaining bits of this word
+        // as a pointer into the message (in octets)
+        cursor_set(cur, orig_ix);
+        int ptr = c0_ntohs(read_bits(cur, data, 16) & 0xFFFF);
+        //@assert ((ptr & 0xc000) == 0xc000);
+        ptr &= 0x3fff;
 

### 2017-12-13 paper.tex
add link to github (feel free to move)

@@
 \begin{abstract}
-  We have implemented a \Kwd{DNS} server in the \Kwd{C0} language, and
-  evaluated it for both safety and performance.
+  We have implemented a \Kwd{DNS} server in the \Kwd{C0} language
+  called \Kwd{c0ns},\footnote{Source code available on GitHub:
+    \url{http://www.github.com/jonsterling/c0ns.git}.} and evaluated it for
+  both safety and performance.
 \end{abstract}

### 2017-12-14 paper.tex
Updated with graphs from external sources + citations.

@@
 
-Because of the low throughput of our system, we did not find it useful to test against commercial nameservers, relying instead on previous work such as \cite{fagin2017making}, which suggests that commercial nameservers can handle loads on the order of 20,000 to 80,000 queries per second.
-With such a large magnitude of difference, even if commercial nameservers perform substantially worse on our benchmark than on the benchmarks that others have used, we still have around 1,000-3,000 times less throughput.
+Because of the low throughput of our system, we did not find it useful to test directly against commercial nameservers, relying instead on previous work such as \cite{fagin2017making}, which suggests that commercial authoritative nameservers can handle loads on the order of 20,000 to 80,000 queries per second, while (at least some) commercial recursive nameservers behave well up to at least 4000 queries per second (See Figure~\ref{fig:commercial-perf}).
+With such a large magnitude of difference, even if commercial nameservers perform substantially worse on our benchmark than on the benchmarks that others have used, we still have around 100-200 times less throughput.
+Additionally, considering latency shows that commercial nameservers tend to have consistent latency once under sufficiently high load, where \Kwd{c0ns}' latency continues to increase with load until it fails to return results within a reasonable amount of time.
+
+We believe that both of these performance issues in \Kwd{c0ns} have the same root cause --- that we only have a single thread, and so incoming requests are processed one at a time.
+In particular, new requests are blocked while we wait for network communication with the remote resolver in order to answer old requests, drastically increasing latency the higher the load on the system.
+Due to the single thread, latency and throughput are directly related as long as there are no failures, so this increased latency also manifests itself as decreased throughput.
 
@@
     \caption{Performance of the checked version of \Kwd{c0ns}}\label{fig:checked-perf}
+\end{figure}
+
+\begin{figure}[h]
+    \includegraphics[width=0.9\textwidth]{Fagin_rate.png}
+    \includegraphics[width=0.9\textwidth]{Fagin_latency.png}
+    \caption{Performance of commercial nameservers. Graphs from \cite{fagin2017making}.}\label{fig:commercial-perf}
 \end{figure}

### 2017-12-14 src/testing/datagen.sh
Added data generation script.

@@ -1,0 +1,6 @@
+#!/bin/bash
+for ii in `cat /usr/share/dict/words`; do
+    echo www.$ii.com A
+    echo www.$ii.net A
+    echo www.$ii.org A
+done

### 2017-12-14 paper.tex
Small change to wording in evaluation:performance.

@@
 Because of the low throughput of our system, we did not find it useful to test directly against commercial nameservers, relying instead on previous work such as \cite{fagin2017making}, which suggests that commercial authoritative nameservers can handle loads on the order of 20,000 to 80,000 queries per second, while (at least some) commercial recursive nameservers behave well up to at least 4000 queries per second (See Figure~\ref{fig:commercial-perf}).
-With such a large magnitude of difference, even if commercial nameservers perform substantially worse on our benchmark than on the benchmarks that others have used, we still have around 100-200 times less throughput.
+With such a large magnitude of difference, even if commercial nameservers perform substantially worse on our benchmark than on the benchmarks that others have used, we still have around 100-200 times less throughput than the maximum shown on these graphs.
 Additionally, considering latency shows that commercial nameservers tend to have consistent latency once under sufficiently high load, where \Kwd{c0ns}' latency continues to increase with load until it fails to return results within a reasonable amount of time.
